---
title: "QC"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
    code-overflow: wrap
execute: 
  cache: true
  warning: false
  error: false
---

## results from zUMI parameter

Use zUMI default.
```
if [ ! -f $out_dir/*_Aligned.sortedByCoord.out.bam ]; then
    # use new 2pass mode of STAR
    STAR --twopassMode Basic \
        --genomeDir $GENOME_DIR \
        --readFilesIn ${R1} ${R2} \
        --runThreadN ${ncpu} \
        --sjdbGTFfile ${GTF} \
        --sjdbOverhang ${sjdbOverhang} \
        --readFilesCommand zcat \
        --outSAMmultNmax 1 \
        --outFilterMultimapNmax 50 \
        --outSAMtype BAM SortedByCoordinate \
        --outSAMattrRGline ID:${sample_ID} LB:library PL:illumina PU:machine SM:${sample_ID} \
        --outFileNamePrefix $out_dir/${sample_ID}_ 

    rm -rf $out_dir/*__STAR*    
	samtools index $out_dir/${sample_ID}_Aligned.sortedByCoord.out.bam
fi
```
```{r, fig.width=10, dev=c('png', 'pdf')}
library(data.table)
library(tidyverse)
library(RColorBrewer)
library(ggsci)

## star summary

star <- fread('../results/multiqc_wells/multiqc_report_data/star_alignment_plot.txt')

# 
colnames(star) <- c('Sample',
  "Uniquely_mapped",
  "Mapped_to_multiple_loci",
  "Mapped_to_too_many_loci",
  # "Unmapped_too_many_mismatches",
  "Unmapped_too_short",
  "Unmapped_other"
)

align_col <- pal_npg(palette = c("nrc"), alpha = 1)(6)
names(align_col) <- colnames(star)[-1]

star <- star %>%
  # 1) Create a Total_Count column
  mutate(Total_Count = rowSums(across(
    c(Uniquely_mapped,
      Mapped_to_multiple_loci,
      Mapped_to_too_many_loci,
      # Unmapped_too_many_mismatches,
      Unmapped_too_short,
      Unmapped_other))),
  Sample = sub("^((?:[^_]+_){3}[^_]+).*", "\\1", Sample)) %>%
  # 2) Create percentage columns for each category
  mutate(across(
    c(Uniquely_mapped,
      Mapped_to_multiple_loci,
      Mapped_to_too_many_loci,
      # Unmapped_too_many_mismatches,
      Unmapped_too_short,
      Unmapped_other),
    ~ (. / Total_Count) * 100,
    .names = "{.col}_pct"
  )) %>%
  separate(Sample, c('Plate', 'Source', 'Treatment', 'Well'), remove = F) %>%
  mutate(is.empty = str_detect(Well, 'A03|C12|F01|H08'))
```

### Summary of STAR aligner
Number of aligned reads in each category
```{r, fig.width=10, dev=c('png', 'pdf')}
star[, c(1:11, 17)] %>%
  pivot_longer(6:11) %>%
  mutate(name = factor(name, levels = c('Total_Count', "Uniquely_mapped",
                                        "Mapped_to_multiple_loci",
                                        "Mapped_to_too_many_loci",
                                        "Unmapped_too_short",
                                        # "Unmapped_too_many_mismatches",
                                        "Unmapped_other"))) %>%
  ggplot(aes(x = Plate, y = value, fill = is.empty, color = is.empty)) +
  geom_violin(width = 1.4, alpha = 0.5, position = position_dodge(width = 1)) +
  geom_boxplot(width = 0.1, alpha = 0.5, position = position_dodge(width = 1)) +
  facet_wrap(.~name, scales = 'free_y', ncol = 1) +
  ylab('Number of reads') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

star[, c(1:10, 17)] %>%
  pivot_longer(6:10) %>%
  mutate(name = factor(name, levels = rev(c("Uniquely_mapped",
                                        "Mapped_to_multiple_loci",
                                        "Mapped_to_too_many_loci",
                                        "Unmapped_too_short",
                                        # "Unmapped_too_many_mismatches",
                                        "Unmapped_other")))) %>%
  ggplot(aes(x = Plate, y = value, fill = name
  )) +
  geom_col(position = 'stack') +
  ylab('Number of reads') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_manual(values = align_col)


star[, c(1:10, 17)] %>%
  pivot_longer(6:10) %>%
  mutate(name = factor(name, levels = rev(c("Uniquely_mapped",
                                        "Mapped_to_multiple_loci",
                                        "Mapped_to_too_many_loci",
                                        "Unmapped_too_short",
                                        # "Unmapped_too_many_mismatches",
                                        "Unmapped_other")))) %>%
  ggplot(aes(x = Plate, y = value, fill = name
  )) +
  geom_col(position = 'fill') +
  ylab('Proportion of reads') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_manual(values = align_col)
```


```
GTF="/vast/projects/lab_davidson/yan.a/ref/gencode/gencode.v44.annotation.gtf"

ncpu=48

# 5. featurecount with deduplicated bam

featureCounts -a $GTF -o unstranded_counts.txt -p -T $ncpu  ../align/*/*_Aligned.sortedByCoord.out.bam
```
This default setting means
```
||             Output file : unstranded_counts.txt                            ||
||                 Summary : unstranded_counts.txt.summary                    ||
||              Paired-end : yes                                              ||
||        Count read pairs : no                                               ||
||              Annotation : gencode.v44.annotation.gtf (GTF)                 ||
||      Dir for temp files : ./                                               ||
||                                                                            ||
||                 Threads : 48                                               ||
||                   Level : meta-feature level                               ||
||      Multimapping reads : not counted                                      ||
|| Multi-overlapping reads : not counted                                      ||
||   Min overlapping bases : 1      
```
Because the BAM files was not mark duplicates, so cannot see duplicate reads. I think the duplicate removal will be different from bulk RNAseq which simply look at read mapping coordinates.  
```{r, fig.width=10, dev=c('png', 'pdf')}
## featurecount

fc <- fread('../results/multiqc_wells/multiqc_report_data/featureCounts_assignment_plot.txt')
fc <- fc[,-3] # remove unmapped counts

colnames(fc) <- c('Sample',
                    "Assigned",
                    "Unassigned_Multi_Mapping",
                    "Unassigned_No_Features",
                    "Unassigned_Ambiguity")

fc_col <- pal_npg(palette = c("nrc"), alpha = 1)(4)
names(fc_col) <- colnames(fc)[-1]

fc <- fc %>%
  # 1) Create a Total_Count column
  mutate(Total_Count = rowSums(across(
    c(Assigned,
      Unassigned_Multi_Mapping,
      Unassigned_No_Features,
      Unassigned_Ambiguity))),
    Sample = sub("^((?:[^_]+_){3}[^_]+).*", "\\1", Sample)) %>%
  # 2) Create percentage columns for each category
  mutate(across(
    c(Assigned,
      Unassigned_Multi_Mapping,
      Unassigned_No_Features,
      Unassigned_Ambiguity),
    ~ (. / Total_Count) * 100,
    .names = "{.col}_pct"
  )) %>%
  separate(Sample, c('Plate', 'Source', 'Treatment', 'Well'), remove = F) %>%
  mutate(is.empty = str_detect(Well, 'A03|C12|F01|H08'))

```


### Summary of featurecount
Number of counted reads in each category
```{r, fig.width=10, dev=c('png', 'pdf')}
fc[, c(1:10, 15)] %>%
  pivot_longer(6:10) %>%
  mutate(name = factor(name, levels = c('Total_Count', "Assigned",
                                        "Unassigned_Multi_Mapping",
                                        "Unassigned_No_Features",
                                        "Unassigned_Ambiguity"))) %>%
  ggplot(aes(x = Plate, y = value, fill = is.empty, color = is.empty)) +
  geom_violin(width = 1.4, alpha = 0.5, position = position_dodge(width = 1)) +
  geom_boxplot(width = 0.1, alpha = 0.5, position = position_dodge(width = 1)) +
  facet_wrap(.~name, scales = 'free_y', ncol = 1) +
  ylab('Number of reads') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

fc[, c(1:10, 15)] %>%
  pivot_longer(6:9) %>%
  mutate(name = factor(name, levels = rev(c("Assigned",
                                        "Unassigned_Multi_Mapping",
                                        "Unassigned_No_Features",
                                        "Unassigned_Ambiguity")))) %>%
  ggplot(aes(x = Plate, y = value, fill = name
  )) +
  geom_col(position = 'stack') +
  ylab('Number of reads') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_manual(values = fc_col)

fc[, c(1:10, 15)] %>%
  pivot_longer(6:9) %>%
  mutate(name = factor(name, levels = rev(c("Assigned",
                                        "Unassigned_Multi_Mapping",
                                        "Unassigned_No_Features",
                                        "Unassigned_Ambiguity")))) %>%
  ggplot(aes(x = Plate, y = value, fill = name
  )) +
  geom_col(position = 'fill') +
  ylab('Proportion of reads') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_manual(values = fc_col)

```

### Number of exonic reads based on qualimap

```{r, fig.width=10, dev=c('png', 'pdf')}
qual_map <- fread('../results/multiqc_merged/multiqc_report_data/qualimap_rnaseq_genome_results.txt') %>%
  # 1) Create a Total_Count column
  mutate(Total_Count = rowSums(across(
    c(reads_aligned_exonic,
      reads_aligned_intronic,
      reads_aligned_intergenic,
      reads_aligned_overlapping_exon)))) %>%
  # 2) Create percentage columns for each category
  mutate(across(
    c(reads_aligned_exonic,
      reads_aligned_intronic,
      reads_aligned_intergenic,
      reads_aligned_overlapping_exon),
    ~ (. / Total_Count) * 100,
    .names = "{.col}_pct"
  )) 
  
qual_map %>% 
  mutate(methods = 'zUMI', Sample = sub("^([^_]+_[^_]+).*", "\\1", Sample)) %>%
  pivot_longer(c(9:12), names_to = 'Genomic_feature', values_to = 'Number_assigned') %>%
  mutate(Genomic_feature = factor(Genomic_feature, 
                                  levels = rev(c("reads_aligned_exonic","reads_aligned_intronic","reads_aligned_intergenic","reads_aligned_overlapping_exon")),
                                  labels = rev(c('exonic','intronic','intergenic','overlapping_exon')))) %>%
  unite("group", c(Sample, methods), remove = F) %>%
  ggplot(aes(x = group, y = Number_assigned, fill = Genomic_feature)) +
  geom_col(position = 'stack') +
  ylab('Number of reads') + 
  # facet_wrap(.~methods, scales = 'free_y', ncol = 1) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  

qual_map %>% 
  mutate(methods = 'zUMI', Sample = sub("^([^_]+_[^_]+).*", "\\1", Sample)) %>%
  pivot_longer(c(9:12), names_to = 'Genomic_feature', values_to = 'Number_assigned') %>%
  mutate(Genomic_feature = factor(Genomic_feature, 
                                  levels = rev(c("reads_aligned_exonic","reads_aligned_intronic","reads_aligned_intergenic","reads_aligned_overlapping_exon")),
                                  labels = rev(c('exonic','intronic','intergenic','overlapping_exon')))) %>%
  unite("group", c(Sample, methods), remove = F) %>%
  ggplot(aes(x = group, y = Number_assigned, fill = Genomic_feature)) +
  geom_col(position = 'fill') +
  ylab('Proportion of reads') + 
  # facet_wrap(.~methods, scales = 'free_y', ncol = 1) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


```