---
title: "SCIMETAR QC using zUMI counts - PD7151"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
    code-overflow: wrap
execute: 
  cache: true
  warning: false
  error: false
---
  
## load package and data
  
Use the same code as Forrest (with minor modification such as row and col names modification).  
Use zUMI parameter generated count matrix and gencode v44 annotation.  

```{r, dev=c('png', 'pdf')}
library(tidyverse)
library(edgeR)
library(data.table)
library(magrittr)
library(Seurat)
library(plyranges)

counts.path <- "../results/gene_counts/collective_gene_counts.txt"
novaseq.counts <- fread(counts.path)[,-c(2:6)] %>% 
  tibble::column_to_rownames('Geneid') %>% 
  as.matrix()

# replace rowname to gene name
# gencode_gtf <- plyranges::read_gff('https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_44/gencode.v44.annotation.gtf.gz')
gencode_gtf <- readRDS('/home/users/allstaff/yan.a/lab_davidson/yan.a/ref/gencode/gencodev44.rds')

genes <- gencode_gtf %>% plyranges::filter(type=='gene') %>%
  plyranges::select(gene_id, gene_name) %>% data.frame()

stopifnot(all.equal(rownames(novaseq.counts), genes$gene_id))

rownames(novaseq.counts) <- genes$gene_name %>% make.unique()

# replace _ with - for seurat later
rownames(novaseq.counts) <- str_replace_all(rownames(novaseq.counts), '_', '-')
colnames(novaseq.counts) <- stringr::str_extract(colnames(novaseq.counts), "^[^-]+")

## compute FPKM
fc_genelength <- fread(counts.path)$Length
names(fc_genelength) <- fread(counts.path)$Geneid

# saveRDS(fc_genelength, 'fc_genelength.rds')

rpkm <- rpkm(novaseq.counts, 
             gene.length = fc_genelength)
rpkm <- rpkm[!(matrixStats::rowSums2(rpkm) == 0), ]

sample.df <- readxl::read_xlsx('data/PD7151 cycle 4 SCIMETAR Plates cross-indexed 20260115.xlsx') %>%
  dplyr::rename(Plate = "Plate...3", Well_ID = "Well_ID...4") %>%
  mutate(Plate = paste0('P', str_pad(Plate, width = 2, pad = "0")))

sample.df <- data.frame(Cell_ID = colnames(novaseq.counts)) %>%
  separate(Cell_ID, c('Plate', 'Source', NA, 'Well_ID'), remove = F) %>% # the middle of the rnaseq id is wrong
  left_join(sample.df, by = c('Plate', 'Well_ID')) %>%
  unite('Plate_Well', Plate, Well_ID, remove = F) %>% 
  mutate(empty_well = Is_Empty,
         Treatment = Cell_Type) # add a dummy treatment which is same as celltype

rownames(sample.df) <- sample.df$Cell_ID

```

## create Seurat object 
Seurat object is more flexible and easy integrate with edgeR later using Seurat2PB function.  

```{r, dev=c('png', 'pdf')}
ss3 <- CreateSeuratObject(counts = rpkm, 
                          assay = 'RNA', 
                          meta.data = sample.df)

# use RPKM to calculate MT/RB/HB percentage
ss3[["percent.mt"]] <- PercentageFeatureSet(ss3, pattern = "^MT-")
ss3[["percent.rb"]] <- PercentageFeatureSet(ss3, pattern = "^RP[SL]")
ss3[["percent.hb"]] <- PercentageFeatureSet(ss3, pattern = "^HB[AB]")

```

## QC (empty well vs non empty)
We would like to see how the empty well behaves in those features.   
So we can use similar cut-off to remove the non-empty wells if they look like empty wells in those features.

```{r, dev=c('png', 'pdf')}
# nCounts_RNA become meaningless here
VlnPlot(ss3, features = c("nFeature_RNA", "percent.mt", "percent.rb", "percent.hb"), 
        split.by = 'empty_well', group.by = 'Source',
        ncol = 2)

VlnPlot(ss3, features = c("nFeature_RNA", "percent.mt", "percent.rb", "percent.hb"), 
        split.by = 'empty_well', group.by = 'Treatment',
        ncol = 2)

FeatureScatter(ss3, feature1 = "nFeature_RNA", feature2 = "percent.mt", 
               split.by = 'empty_well', group.by = 'Source')
FeatureScatter(ss3, feature1 = "nFeature_RNA", feature2 = "percent.rb", 
               split.by = 'empty_well', group.by = 'Source')
FeatureScatter(ss3, feature1 = "nFeature_RNA", feature2 = "percent.hb", 
               split.by = 'empty_well', group.by = 'Source')
```


```{r, dev=c('png', 'pdf')}

# log1p normalization and calculate cell cycle score, this replace the NormalizeData step
ss3[['RNA']]$data <- log1p(ss3[['RNA']]$counts)

# FindVariableFeatures use counts slots
ss3 <- FindVariableFeatures(ss3, 
                            selection.method = "vst", 
                            nfeatures = 2000)
ss3 <- ScaleData(ss3, features = rownames(ss3))

ss3 <- CellCycleScoring(object = ss3, 
                        g2m.features = cc.genes.updated.2019$g2m.genes,
                        s.features = cc.genes.updated.2019$s.genes)

VlnPlot(ss3, features = c('PCNA','MKI67','TOP2A','MCM6'),
        split.by = 'empty_well', cols = 2, group.by = 'Source')

```


## Identify the 10 most highly variable genes
```{r, dev=c('png', 'pdf')}
top10 <- head(VariableFeatures(ss3), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(ss3)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2

```

## PCA plot and loading in PC1 and PC2
```{r, dev=c('png', 'pdf')}

ss3 <- RunPCA(ss3, features = VariableFeatures(object = ss3))

ElbowPlot(ss3)

VizDimLoadings(ss3, dims = 1:2, reduction = "pca")

DimPlot(ss3, reduction = "pca") 

```

## UMAP
```{r, dev=c('png', 'pdf')}
ss3 <- FindNeighbors(ss3, dims = 1:10)
ss3 <- FindClusters(ss3, resolution = 0.5)

ss3 <- RunUMAP(ss3, dims = 1:10, n.neighbors = 30)
DimPlot(ss3, reduction = "umap")

DimPlot(ss3, reduction = "umap", group.by = 'empty_well')

DimPlot(ss3, reduction = "umap", group.by = 'Phase')
DimPlot(ss3, reduction = "umap", group.by = 'Source')
DimPlot(ss3, reduction = "umap", group.by = 'Treatment')

DimPlot(ss3, reduction = "umap", group.by = 'Plate', split.by = 'Treatment')
```


```{r, fig.height=15, dev=c('png', 'pdf')}
FeaturePlot(ss3, split.by = 'empty_well',
            features = c("CD34", "TET2", "GAPDH", "CD33")) & theme(legend.position = "right")

FeaturePlot(ss3, split.by = 'empty_well',
            features = c("PTPRC","HBB", 'HBA1','HBA2')) & theme(legend.position = "right")

# nCount is meaningless for RPKM
FeaturePlot(ss3, split.by = 'empty_well',
            features = c("nFeature_RNA", "percent.mt", "percent.rb", "percent.hb")) & theme(legend.position = "right")

```


```{r, dev=c('png', 'pdf')}
# Calculate cell counts per cluster
cluster_counts <- data.frame(cluster = ss3$seurat_clusters, 
                             Source = ss3$Source,
                             Treatment = ss3$Treatment,
                             empty = ss3$empty_well, 
                             Plate = ss3$Plate)

# Plot using a barplot
DimPlot(ss3, split.by = 'empty_well',
        reduction = "umap", 
        group.by = 'Source') & theme(legend.position = "right")

ggplot(cluster_counts, aes(x = cluster)) +
  geom_bar(aes(fill = Source), position = "fill") +
  labs(x = "Cluster", y = "Proportion") 

DimPlot(ss3, split.by = 'empty_well',
        reduction = "umap", 
        group.by = 'Treatment') & theme(legend.position = "right")

ggplot(cluster_counts, aes(x = cluster)) +
  geom_bar(aes(fill = Treatment), position = "fill") +
  labs(x = "Cluster", y = "Proportion") 

ggplot(cluster_counts, aes(x = cluster)) +
  geom_bar(aes(fill = Plate), position = "fill") +
  labs(x = "Cluster", y = "Proportion") 

```


```{r, dev=c('png', 'pdf')}

saveRDS(ss3, 'output/seurat_zumi_rpkm_unfilt2.rds')

```