---
title: "Amplicon summary"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
    code-overflow: wrap
    number-sections: true
    number-depth: 3
execute:
  warning: false
  error: false
  cache: true
---

```{r, fig.width=15, fig.height=10, dev=c('png', 'pdf')}

library(ggplot2)
library(tidyverse)
library(data.table)
library(readxl)
library(ggpubr)
library(future.apply)
source('../script/summarize_mutation_file.R')

```

## Load new demux and mut calling results

Now we first demultiplex using the barcode and still keep R1 and R2 separately (reads cannot demux are discarded).

1\. In round 1: For R1 and R2, we try to do mutation calling separately and record if mutation is found in R1 and R2-reverse complement using an exact match 5bp centred around the mutation sites. Reads are marked with R1 or R2rc.\
2. In round 2: For the R1 and R2 that did not pass round 1 (unmatched), we search again only for SRSF, now we use 10bp flanking the mutation site, and allow for 10% mismatch (equivalent to 1 bp). Reads are marked with R1_2.

\
Then we count:

1\. SRSF2 as reads with mutations or WT in R2 only, or both R1/R2rc.

2\. TET2a as reads with mutations or WT in R1 only or both R1/R2rc.

3\. TET2b as reads with mutations or WT in R2 only or both R1/R2rc.

```{r, fig.width=15, fig.height=10, dev=c('png', 'pdf')}

files.mut1 <- list.files('../amplicon/run1/', '_merged_all_hits.tsv', recursive = T, full.names = T)

plan(multisession, workers = 10)

mut1.summary <- future_lapply(files.mut1, 
               summarize_mutation_file)

mutation_stats1 <- future_lapply(1:length(files.mut1), function(x) {
  mut1.summary[[x]] %>% 
  mutate(file = basename(files.mut1[x])) %>% 
  separate(file, c("Plate", 'Well')) %>%
  separate(mutation, c("Gene", "genotype"), remove = F) %>%
  dplyr::rename(interaction = mutation) %>%
  filter(
    (str_detect(interaction, "tet2a") & read_types %in% c("R1_R2rc", "R1")) |
    (str_detect(interaction, "tet2b") & read_types %in% c("R1_R2rc", "R2rc")) |
    (str_detect(interaction, "srsf2") & (read_types %in% c("R1_R2rc", "R2rc")))
  ) %>%
  group_by(interaction, Plate, Well) %>%
  summarize(Count = sum(read_count, na.rm = TRUE))
}) %>% rbindlist() 
```

```{r, fig.width=15, fig.height=10, dev=c('png', 'pdf')}

files.mut2 <- list.files('../amplicon/run2/', '_merged_all_hits.tsv', recursive = T, full.names = T)
mut2.summary <- future_lapply(files.mut2, 
               summarize_mutation_file)

mutation_stats2 <- future_lapply(1:length(files.mut2), function(x) {
  mut2.summary[[x]] %>% 
  mutate(file = basename(files.mut2[x])) %>% 
  separate(file, c("Plate", 'Well')) %>%
  separate(mutation, c("Gene", "genotype"), remove = F) %>%
  dplyr::rename(interaction = mutation) %>%
  filter(
    (str_detect(interaction, "tet2a") & read_types %in% c("R1_R2rc", "R1")) |
    (str_detect(interaction, "tet2b") & read_types %in% c("R1_R2rc", "R2rc")) |
    (str_detect(interaction, "srsf2") & (read_types %in% c("R1_R2rc", "R2rc") ))
  ) %>%
  group_by(interaction, Plate, Well) %>%
  summarize(Count = sum(read_count, na.rm = TRUE))
}) %>% rbindlist() 
```

```{r, fig.width=15, fig.height=10, dev=c('png', 'pdf')}
list(mutation_stats1 %>% separate(interaction, c('Gene','genotype'), remove = F) %>% mutate(version = 'Run1'),
     mutation_stats2 %>% separate(interaction, c('Gene','genotype'), remove = F) %>% mutate(version = 'Run2')) %>%
  rbindlist(use.names=TRUE) %>%
  filter(interaction != 'unmatched_unmatched') %>%
  ggplot(aes(x = interaction, y = Count, fill = version)) +
  geom_col(position = "dodge") +
  facet_wrap(. ~ Plate, nrow = 6, scales = 'free_y') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


```{r, dev=c('png', 'pdf')}
list(mutation_stats1 %>% separate(interaction, c('Gene','genotype'), remove = F) %>% mutate(version = 'V1'),
     mutation_stats2 %>% separate(interaction, c('Gene','genotype'), remove = F) %>% mutate(version = 'V2')) %>%
  rbindlist(use.names=TRUE) %>%
  filter(interaction %in% c('srsf2_wt','srsf2_mut')) %>% 
  # The 'genotype' column corresponds to your requested 'mutation' facet
  pivot_wider(
    names_from = version,
    values_from = Count,
    id_cols = c(interaction, Gene, genotype, Plate, Well)
  ) %>% 
  # --- 3. Generate the scatter plot with facets and color ---
  ggplot(aes(x = V1, y = V2, color = Plate)) +
  # Add the points
  geom_point(size = 3) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray") +
  # Create a facet for each 'mutation' (genotype)
  facet_wrap(~ genotype) +
  # Add labels and a title for clarity
  labs(
    x = "Run1 Count",
    y = "Run2 Count",
    title = "Scatter Plot of Run1 vs Run2 Counts",
    color = "Plate"
  ) 
```

```{r, dev=c('png', 'pdf')}
write.csv(mutation_stats1, 'output/mutation_stats1.csv', row.names = F)
write.csv(mutation_stats2, 'output/mutation_stats2.csv', row.names = F)

pcr2 <- read.csv('output/LINE1_qPCR_all_corrected.csv')

df_wide2 <- mutation_stats1 %>%
  left_join(mutation_stats2, by = c("interaction",  "Plate",   "Well"),
            suffix = c('.run1', '.run2')) %>%
  replace_na(list(Count.run1 = 0, Count.run2 = 0)) %>% 
  complete(interaction, Plate, Well, fill = list(Count.run1 = 0, Count.run2 = 0)) %>%
  mutate(Count = Count.run1 + Count.run2) %>%
  separate(interaction, c('Gene','genotype')) %>% 
  # mutate(version = 'V2') %>%
  left_join(pcr2, by = c('Plate', 'Well')) %>%
  filter(genotype %in% c("wt", "mut")) %>%   # drop “unmatched” or other genotypes if you only want wt+mut
  select(-Count.run1, -Count.run2) %>% 
  pivot_wider(
    names_from   = genotype,
    values_from  = Count
  )
write.csv(df_wide2, 'output/amplicon_read_num_sum_revise3.csv')

```
