---
title: "SCIMETAR single cell analysis (Flitered by RPKM) add LINE1, FACS and amplicon data"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
    code-overflow: wrap
    number-sections: true
    number-depth: 3
execute: 
  cache: true
  warning: false
  error: false
---

## Load filtered cells 


```{r, dev=c('png', 'pdf')}
library(matrixStats)
library(ggfortify)
library(Seurat)
library(edgeR)
library(tidyverse)
library(DT)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggpubr)
library(EnhancedVolcano)
library(ggh4x)
library(viridis)
library(patchwork)
library(readxl)
library(RColorBrewer)
library(UpSetR)
library(data.table)
library(ggsankey)
library(ggalluvial)
library(ggrepel)
library(scales)
library(ggsci)
library(Azimuth)
```

```{r}
seuratobj <- readRDS('output/seurat_zumi_rpkm_unfilt2.rds')
```

I use nFeature_RNA > 1000 & not empty & remove DEAD cells in FACS annotation.  
There are 1582 cells passed the filter.  
```{r}
## load FACS data
## load LINE1 data
pcr <- read.csv('output/LINE1_qPCR_all_corrected.csv') 

## load amplicond data, new version
## this is generated from the shinyapp https://shinyapp.curtis-lab-analysis.cloud.edu.au/shiny/scimetar_amplicon_genotype/ using output from 4.amplicon_summary_5.qmd
amplicon <- read.csv('output/genotyped_data_2026-01-30.csv') %>%
  dplyr::select(7:16) %>%
  mutate(across(matches("wt_|mut_"), ~ log1p(.x), 
                .names = "{.col}_log1p"))

master <- pcr %>%
  left_join(amplicon, by = 'Plate_Well') %>%
  mutate(
    celltype_merge = factor(celltype,
                                 levels = c("HSC","MPP","LMPP","CMP","CLP","MEP","GMP","other", "empty", NA)),
         group = factor(group, 
                        levels = c("C4D1_C4D1","C4D8_C4D8")))

rownames(master) <- master$Cell_ID
write.csv(master, 'output/master.csv', quote = F, row.names = F)

seuratobj <- AddMetaData(seuratobj, 
                         master)

filtered2 <- subset(seuratobj, subset = nFeature_RNA > 1000 & 
                      empty_well == FALSE & celltype_merge != 'DEAD' & !is.na(celltype_merge))

filtered2@meta.data %>% dplyr::count(Cell_Type, Treatment)
```


```{r}

na_grey   <- "grey85"            # any NA in celltype_merge

# use the Dev Cell colors extracted, same order as the weblink
col_ref <- c(
  HSC   = "#ED6D52",
  MPP1  = "#F5C3CB",
  MPP2  = "#DF87E8",
  LMPP  = "#5DCBCF",
  MLP   = "#4A895B",
  MEP   = "#8E71D5",
  Neu1  = "#F2A93B",
  B_NK1 = "#75FB8C",
  Neu2  = "#C3884D",
  MD    = "#EFE697",
  EBM   = "#FADFB3",
  Mk    = "#4A68DA",
  Ery   = "#97CCE7",
  B_NK2 = "#A6EB99"
)

col_celltype <- c(
  DEV_HSC   = "#ED6D52", FACS_HSC = "#ED6D52", 
  DEV_MPP1  = "#F5C3CB", #FACS_MHSC = "#DF87E8", 
  FACS_MPP = "#F5C3CB",
  DEV_MPP2  = "#DF87E8",
  DEV_MEP   = "#8E71D5", FACS_MEP = "#8E71D5",
  DEV_Ery   = "#97CCE7",
  DEV_Mk    = "#4A68DA",
  DEV_LMPP  = "#5DCBCF", FACS_LMPP = "#5DCBCF", FACS_CMP = "#F2A93B", 
  DEV_MLP   = "#4A895B", FACS_CLP = "#4A895B",
  DEV_B_NK1 = "#75FB8C",
  DEV_B_NK2 = "#A6EB99",
  DEV_Neu1  = "#F2A93B", FACS_GMP = "#C3884D", 
  DEV_Neu2  = "#C3884D",
  DEV_MD    = "#EFE697",
  DEV_EBM   = "#FADFB3"
)

col_celltype <- col_celltype[c(
  "FACS_HSC",'DEV_HSC',
  "FACS_MHSC","FACS_MPP","DEV_MPP1","DEV_MPP2",
  "FACS_CMP","FACS_LMPP",'DEV_LMPP',
  "FACS_GMP", "DEV_Neu1","FACS_MEP",'DEV_MEP',"FACS_CLP","DEV_MLP",
  "DEV_Neu2","DEV_B_NK1",
  "DEV_MD","DEV_EBM","DEV_Ery","DEV_Mk")]

# col_facs <- setNames(pal_npg("nrc")(10)[1:9], filtered2@meta.data$celltype_merge %>% factor %>% levels)
# col_facs["other"] <- "grey50"        # explicit grey for “other”
col_facs <- c(
  HSC = "#ED6D52",
  #MHSC = "#DF87E8", 
  MPP = "#F5C3CB",
  LMPP = "#5DCBCF",
  CMP = "#F2A93B", 
  CLP = "#4A895B",
  MEP = "#8E71D5",
  GMP = "#C3884D",
  other = "grey50"
)

# col_group <- setNames(brewer.pal(12, "Paired")[c(3:6)], filtered2@meta.data$group %>% factor %>% levels)
col_group <- c("C4D1_C4D1" = "#80cdc1", 
               "C4D8_C4D8" = "#018571"
               # "C4D8_C4D8" = "#dfc27d",
               # 'Primary_Patient_AZA' = "#a6611a"
               )

col_cluster <- setNames(brewer.pal(12, "Set3")[c(1,3:8,10,11)], 0:8)

col_treatment <- setNames(brewer.pal(2, "Set2")[1:2], c('C4D1','C4D8'))

col_source <- setNames(c("#018571" , "#a6611a"), c('GR', 'Primary_Patient'))

col_phase <- c(G1='#663399', G2M='#E66100', S='#2A934E')
```

## Top variable features and PCA/UMAP plot
Top 10 most variable genes still have HB genes.  
After clustering, there are 5 major clusters.

```{r, dev=c('png', 'pdf')}
filtered2 <- FindVariableFeatures(filtered2, 
                                  selection.method = "vst", 
                                  nfeatures = 2000)

filtered2 <- ScaleData(filtered2, features = rownames(filtered2))

filtered2 <- RunPCA(filtered2, features = VariableFeatures(object = filtered2))
VizDimLoadings(filtered2, dims = 1:2, reduction = "pca")

DimPlot(filtered2, reduction = "pca") 

filtered2 <- CellCycleScoring(object = filtered2, 
                              g2m.features = cc.genes.updated.2019$g2m.genes,
                              s.features = cc.genes.updated.2019$s.genes)

filtered2 <- FindNeighbors(filtered2, dims = 1:20)
filtered2 <- FindClusters(filtered2, resolution = 1)

filtered2 <- RunUMAP(filtered2, dims = 1:20, 
                     n.neighbors = 50)

DimPlot(filtered2, reduction = "umap", label = T, cols = col_cluster) +
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )

DimPlot(filtered2, reduction = "umap", 
        split.by = 'seurat_clusters',
        cols = col_cluster) 

DimPlot(filtered2, reduction = "umap",  group.by = 'Phase',
        cols = col_phase) + 
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )

DoHeatmap(filtered2, 
          features = c(cc.genes.updated.2019$g2m.genes, cc.genes.updated.2019$s.genes),
          group.by = 'Phase', group.colors = col_phase,
          angle = 90) 
```

### UMAP plot by plate 
to examine batch effect from plate
```{r, dev=c('png', 'pdf')}
#| fig-cap: "UMAP plot colored by different plate"
#| layout-ncol: 2

DimPlot(filtered2, reduction = "umap", group.by = 'Plate', 
        cols = c(brewer.pal(9, 'Set3'), rep('grey90', 11)))  +
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )

DimPlot(filtered2, reduction = "umap", group.by = 'Plate', 
        cols = c(rep('grey90', 9), brewer.pal(11, 'Set3')))  +
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )

```


### UMAP plot by treatment and cell types
```{r, dev=c('png', 'pdf')}
#| fig-cap: "UMAP plot colored by different grouping"
#| layout-ncol: 3

DimPlot(filtered2, reduction = "umap",  group.by = 'Treatment',
        cols = col_treatment) +
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )

DimPlot(filtered2, reduction = "umap",  group.by = 'group',
        cols = col_group) +
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )

DimPlot(filtered2, reduction = "umap",  group.by = 'celltype_merge',
        cols = col_facs,   na.value = na_grey) +  
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )
```


```{r, dev=c('png', 'pdf'), fig.width=7, fig.width=7}

for (group in names(col_group)) {
  
  # Create highlight column
  filtered2$highlighted_group <- ifelse(filtered2$group == group, group, NA)
  
  # Define colours
  highlight_cols <- c(col_group[group], "grey95")
  names(highlight_cols) <- c(group, "NA")
  
  # Reorder cells: grey (NA) first, highlighted cells last
  ordered_cells <- filtered2@meta.data %>%
    mutate(cell = rownames(.)) %>%
    arrange(!is.na(highlighted_group)) %>%   # NA first, highlighted last
    pull(cell)
  
  # Plot with ordered cells
  p <- DimPlot(
    filtered2,
    reduction = "umap",
    group.by = "highlighted_group",
    cols = highlight_cols,
    na.value = "grey95",
    cells = ordered_cells
  ) +
    ggtitle(paste("UMAP with", group, "Highlighted")) +
    plot_layout(ncol = 1, nrow = 1,
                widths = unit(3, "in"),
                heights = unit(3, "in"))
  
  print(p)
}


```

```{r, dev=c('png', 'pdf'), fig.width=20, fig.height=6}

DimPlot(filtered2, reduction = "umap", 
        split.by = 'group', 
        cols = col_group, group.by = 'group') +
  plot_layout(ncol = 1, nrow = 1,     
                        widths = unit(6, "in"),   
                        heights = unit(3, "in")) 

DimPlot(filtered2, reduction = "umap", 
        split.by = 'group', 
        cols = col_cluster) +
    plot_layout(ncol = 1, nrow = 1,     
                        widths = unit(6, "in"),   
                        heights = unit(3, "in")) 

```

```{r, dev=c('png', 'pdf')}
#| fig-cap: "Bar plot colored by different grouping"
#| layout-ncol: 3

ggplot(filtered2@meta.data, aes(x = seurat_clusters)) +
  geom_bar(aes(fill = Treatment), position = "fill") +
    scale_fill_manual(
    values   = col_treatment,
    name     = "Treatment"
  ) +
  labs(x = "Cluster", y = "Proportion") +
  force_panelsizes(rows = unit(4, 'inch'), cols = unit(3, 'inch')) 

ggplot(filtered2@meta.data, aes(x = seurat_clusters)) +
  geom_bar(aes(fill = group), position = "fill") +
  scale_fill_manual(
    values   = col_group,
    name     = "Source_Treatment"
  ) +
  labs(x = "Cluster", y = "Proportion") +
  force_panelsizes(rows = unit(4, 'inch'), cols = unit(3, 'inch')) 

ggplot(filtered2@meta.data, aes(x = seurat_clusters)) +
  geom_bar(aes(fill = celltype_merge), position = "fill") +
  scale_fill_manual(
    values   = col_facs,
    na.value = na_grey,
    name     = "FACS Immunophenotype"
  ) +
  labs(x = "Cluster", y = "Proportion") +
  force_panelsizes(rows = unit(4, 'inch'), cols = unit(3, 'inch')) 

# 0: HSC/MHSC
# 7: lymphoid
# 7/8: CB DMSO

```

### AZA-induced Seurat cluster expansion/shrinking
```{r, dev=c('png', 'pdf'), fig.width=8, fig.height=6}

ggplot(filtered2@meta.data, aes(x = group)) +
  geom_bar(aes(fill = seurat_clusters), position = "fill") +
    scale_fill_manual(
    values   = col_cluster,
    na.value = na_grey,
    name     = "Seurat Clusters"
  ) +
  facet_grid(. ~ Cell_Type, scales = 'free_x') +
  labs(x = "Group", y = "Proportion") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  force_panelsizes(rows = unit(4, 'inch'), cols = unit(3, 'inch')) 

props <- filtered2@meta.data %>%
  count(group, seurat_clusters, name = "n", .drop = FALSE) %>%
  group_by(group) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  separate(group, c(NA, 'Cell_Type', 'Treatment'), remove = F)

ggplot(props, aes(x = group, y = prop, group = seurat_clusters, color = seurat_clusters)) +
  geom_line() +
  geom_point(size = 3) +
  geom_text_repel(
    aes(label = scales::percent(prop, accuracy = 0.1)),
    size = 3, show.legend = FALSE
  ) +
  facet_grid(. ~ Cell_Type, scales = 'free_x') +
  labs(x = NULL, y = "Proportion") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  force_panelsizes(rows = unit(4, 'inch'), cols = unit(3, 'inch')) +
  scale_color_manual(
    values   = col_cluster,
    na.value = na_grey,
    name     = "Seurat Clusters"
  ) 

```


### AZA-induced FACS cluster expansion/shrinking
```{r, dev=c('png', 'pdf'), fig.width=8, fig.height=6}

ggplot(filtered2@meta.data, aes(x = group)) +
  geom_bar(aes(fill = celltype_merge), position = "fill") +
    scale_fill_manual(
    values   = col_facs,
    na.value = na_grey,
    name     = "FACS"
  ) +
  facet_grid(. ~ Cell_Type, scales = 'free_x') +
  labs(x = "Group", y = "Proportion") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  force_panelsizes(rows = unit(4, 'inch'), cols = unit(3, 'inch')) 

props <- filtered2@meta.data %>%
  count(group, celltype_merge, name = "n", .drop = FALSE) %>%
  group_by(group) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  separate(group, c(NA, 'Cell_Type', 'Treatment'), remove = F)

ggplot(props, aes(x = group, y = prop, group = celltype_merge, color = celltype_merge)) +
  geom_line() +
  geom_point(size = 3) +
  geom_text_repel(
    aes(label = scales::percent(prop, accuracy = 0.1)),
    size = 3, show.legend = FALSE
  ) +
  facet_grid(. ~ Cell_Type, scales = 'free_x') +
  labs(x = NULL, y = "Proportion") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  force_panelsizes(rows = unit(4, 'inch'), cols = unit(3, 'inch')) +
  scale_color_manual(
    values   = col_facs,
    na.value = na_grey,
    name     = "FACS"
  ) 

ggplot(props %>%
         filter(!(celltype_merge %in% c('DEAD','other'))), 
       aes(x = celltype_merge, y = prop, fill = group)) +
  geom_col(position = position_dodge(width = 0.8, preserve = "single"), 
           width = 0.8) +
  labs(x = "Cell type", y = "Proportion", fill = "Group") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(.~Cell_Type, scales = "free_x", space = "free_x") +
  scale_fill_manual(
    values   = col_group,
    na.value = na_grey,
    name     = "Group"
  )

```


### Find cluster marker gene
Top 20 markers for all clusters.  
```{r, dev=c('png', 'pdf'), fig.width=15, fig.height=5}

cluster.markers <- FindAllMarkers(filtered2, only.pos = TRUE)

# Create an interactive datatable, show only top 20 (by p value) positive markers in each cluster
datatable(cluster.markers[, c(6,3,4,2,1,5,7)] %>%
            group_by(cluster) %>% 
            arrange(p_val) %>% 
            slice_head(n = 20) %>% 
            ungroup(), 
          filter = "top", caption = 'Top 20 DEG in each cluster',
          options = list(pageLength = 20, autoWidth = TRUE)) %>%
  formatSignif(columns = 2:6, digits = 3)

# selected marker based on pct, logFC and p value
selected_marker <- cluster.markers %>%
  group_by(cluster) %>% 
  arrange(p_val) %>% 
  slice_head(n = 3) %>% 
  ungroup() %>% pull(gene)

DotPlot(filtered2, features = selected_marker %>% unique())  + 
  RotatedAxis() +
  force_panelsizes(rows = unit(3, 'inch'), cols = unit(length(selected_marker)/4, 'inch')) +
  scale_color_viridis(direction = 1) +
  ggtitle('Top 3 markers in each cluster')

selected_marker <- cluster.markers %>%
  group_by(cluster) %>% 
  arrange(p_val) %>% 
  filter(pct.1 > 0.5, pct.2 < 0.2, avg_log2FC > 1) %>%
  slice_head(n = 3) %>% 
  ungroup() %>% pull(gene)

DotPlot(filtered2, features = selected_marker %>% unique())  + 
  RotatedAxis() +
  force_panelsizes(rows = unit(3, 'inch'), cols = unit(length(selected_marker)/4, 'inch')) +
  scale_color_viridis(direction = 1) +
  ggtitle('Top 3 markers in each cluster\n(pct.cluster > 0.5, pct.rest < 0.2, avg_log2FC > 1)')

# show the markers from Dev cell paper
marker <- read.table('../../R/data/dev_cell_2022/Cluster_marker.txt', sep = '\t', header = T)

marker <- marker %>% filter(avg_logFC>0 ) %>%
  group_by(cluster) %>%
  arrange(p_val_adj) %>%
  slice_head(n=3) %>% 
  pull(gene) %>%
  unique()

DotPlot(filtered2, features = marker)  + 
  RotatedAxis() +
  force_panelsizes(rows = unit(3, 'inch'), cols = unit(length(marker)/4, 'inch')) +
  scale_color_viridis(direction = 1) +
  ggtitle('Top 3 markers in DevCell subtype from their paper')

```

### GSEA analysis of MSigDb Hallmark
```{r, fig.height=50, fig.width=15, dev=c('png', 'pdf')}
library(msigdbr)
msigdb_sets <- msigdbr(species = "Homo sapiens", category = "H")

hm_sets <- split(x = msigdb_sets$gene_symbol, f = msigdb_sets$gs_name)

# saveRDS(hm_sets, 'data/hallmark_geneset.rds')

filtered2 <- AddModuleScore(filtered2, 
                            features = hm_sets, 
                            name = names(hm_sets))

filtered2@meta.data %>%
  pivot_longer(HALLMARK_ADIPOGENESIS1:HALLMARK_XENOBIOTIC_METABOLISM50, 
               names_to = 'GeneSet', values_to = 'Score') %>%
  mutate(GeneSet = str_remove(GeneSet, 'HALLMARK_|GOBP_')) %>%
  ggplot(aes(x = seurat_clusters, y = Score, col = seurat_clusters)) +
  geom_violin(trim = FALSE, alpha = 0.5) +  # Violin plot with no trimming
  geom_boxplot(width = 0.3, alpha = 0.8) +
  facet_wrap(.~GeneSet, scales = 'free_y', ncol = 4) +
  stat_compare_means() +
  force_panelsizes(rows = unit(2.5, 'inch'), cols = unit(2.5, 'inch')) 

```

## Cell type annotation 

### use Dev Cell 2022 data 
```{r, dev=c('png', 'pdf')}

## run azimuth

filtered2 <- RunAzimuth(
  query = filtered2,
  reference = "ref_azimuth/" 
)

filtered2$predicted.RNA_clusters <- factor(filtered2$predicted.RNA_clusters, 
                                            levels = names(col_ref))

DimPlot(filtered2, reduction = "umap",  group.by = 'predicted.RNA_clusters',
        cols = col_ref) +
  ggtitle("Annotated to Dev Cell using Azimuth") +
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )

ggplot(filtered2@meta.data, aes(x = seurat_clusters)) +
  geom_bar(aes(fill = predicted.RNA_clusters), position = "fill") +
  scale_fill_manual(
    values   = col_ref,
    na.value = na_grey,
    name     = "Azimuth Dev Cell"
  ) +
  labs(x = "Cluster", y = "Proportion") +
  force_panelsizes(rows = unit(4, 'inch'), cols = unit(3, 'inch')) 

## compare dev cell vs facs clusters
filtered2@meta.data %>%
  group_by(Cell_Type, Treatment) %>%
  group_modify(~make_long(.x, celltype_merge, predicted.RNA_clusters)) %>%
  mutate(node = factor(node, 
                       levels = c(
                         rev(names(col_ref)),
                         rev(names(col_facs))) %>% rev() %>% unique() %>% rev() ) 
  ) %>% 
  ggplot(aes(x = x, 
             next_x = next_x, 
             node = node, 
             next_node = next_node,
             fill = node,
             label = node)) +
  geom_sankey(flow.alpha = .6,
              node.color = "gray30") +
  geom_sankey_label(size = 3, color = "white", fill = "gray40") +
  # scale_fill_viridis_d(drop = FALSE) +
  scale_fill_manual(values = c(col_facs, 
                               col_ref
  ))  

saveRDS(filtered2, 'output/filtered2.rds')
filtered2@meta.data %>% write.csv('output/master_filtered.csv')
```


```{r}
sessionInfo()
```
