---
title: "SCIMETAR single cell analysis (Flitered by RPKM) add LINE1, FACS and amplicon data"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
    code-overflow: wrap
    number-sections: true
    number-depth: 3
execute: 
  cache: true
  warning: false
  error: false
---

## Load filtered and annotated seurat object 

```{r, dev=c('png', 'pdf')}
library(matrixStats)
library(ggfortify)
library(Seurat)
library(edgeR)
library(tidyverse)
library(DT)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggpubr)
library(EnhancedVolcano)
library(ggh4x)
library(viridis)
library(patchwork)
library(readxl)
library(RColorBrewer)
library(UpSetR)
library(data.table)
library(ggsankey)
library(ggalluvial)
library(ggrepel)
library(scales)
library(ggsci)
```


```{r}

filtered2 <- readRDS("output/filtered2.rds")

na_grey   <- "grey85"            # any NA in celltype_merge

# use the Dev Cell colors extracted, same order as the weblink
col_ref <- c(
  HSC   = "#ED6D52",
  MPP1  = "#F5C3CB",
  MPP2  = "#DF87E8",
  LMPP  = "#5DCBCF",
  MLP   = "#4A895B",
  MEP   = "#8E71D5",
  Neu1  = "#F2A93B",
  B_NK1 = "#75FB8C",
  Neu2  = "#C3884D",
  MD    = "#EFE697",
  EBM   = "#FADFB3",
  Mk    = "#4A68DA",
  Ery   = "#97CCE7",
  B_NK2 = "#A6EB99"
)

col_celltype <- c(
  DEV_HSC   = "#ED6D52", FACS_HSC = "#ED6D52", 
  DEV_MPP1  = "#F5C3CB", #FACS_MHSC = "#DF87E8", 
  FACS_MPP = "#F5C3CB",
  DEV_MPP2  = "#DF87E8",
  DEV_MEP   = "#8E71D5", FACS_MEP = "#8E71D5",
  DEV_Ery   = "#97CCE7",
  DEV_Mk    = "#4A68DA",
  DEV_LMPP  = "#5DCBCF", FACS_LMPP = "#5DCBCF", FACS_CMP = "#F2A93B", 
  DEV_MLP   = "#4A895B", FACS_CLP = "#4A895B",
  DEV_B_NK1 = "#75FB8C",
  DEV_B_NK2 = "#A6EB99",
  DEV_Neu1  = "#F2A93B", FACS_GMP = "#C3884D", 
  DEV_Neu2  = "#C3884D",
  DEV_MD    = "#EFE697",
  DEV_EBM   = "#FADFB3"
)

col_celltype <- col_celltype[c(
  "FACS_HSC",'DEV_HSC',
  "FACS_MHSC","FACS_MPP","DEV_MPP1","DEV_MPP2",
  "FACS_CMP","FACS_LMPP",'DEV_LMPP',
  "FACS_GMP", "DEV_Neu1","FACS_MEP",'DEV_MEP',"FACS_CLP","DEV_MLP",
  "DEV_Neu2","DEV_B_NK1",
  "DEV_MD","DEV_EBM","DEV_Ery","DEV_Mk")]

# col_facs <- setNames(pal_npg("nrc")(10)[1:9], filtered2@meta.data$celltype_merge %>% factor %>% levels)
# col_facs["other"] <- "grey50"        # explicit grey for “other”
col_facs <- c(
  HSC = "#ED6D52",
  #MHSC = "#DF87E8", 
  MPP = "#F5C3CB",
  LMPP = "#5DCBCF",
  CMP = "#F2A93B", 
  CLP = "#4A895B",
  MEP = "#8E71D5",
  GMP = "#C3884D",
  other = "grey50"
)

# col_group <- setNames(brewer.pal(12, "Paired")[c(3:6)], filtered2@meta.data$group %>% factor %>% levels)
col_group <- c("C4D1_C4D1" = "#80cdc1", 
               "C4D8_C4D8" = "#018571"
               # "C4D8_C4D8" = "#dfc27d",
               # 'Primary_Patient_AZA' = "#a6611a"
               )

col_cluster <- setNames(brewer.pal(12, "Set3")[c(1,3:8,10,11)], 0:8)

col_treatment <- setNames(brewer.pal(2, "Set2")[1:2], c('C4D1','C4D8'))

col_source <- setNames(c("#018571" , "#a6611a"), c('GR', 'Primary_Patient'))

col_phase <- c(G1='#663399', G2M='#E66100', S='#2A934E')

col_genotype <- c(
  "unidentifiable" = '#d9d9d9',
  "wt/wt" = '#253494',
  "mut/wt" = '#dd1c77',
  # "mut/mut" = '#980043',
  "mut/mut" = '#dd1c77'
)

# 2 group genotype
filtered2 <- AddMetaData(filtered2, 
                         metadata = data.frame(genotype_srsf2_2 = str_replace(filtered2$genotype_srsf2, 'mut/mut','mut/wt'),
                                               genotype_tet2a_2 = str_replace(filtered2$genotype_tet2a, 'mut/mut','mut/wt'),
                                               genotype_tet2b_2 = str_replace(filtered2$genotype_tet2b, 'mut/mut','mut/wt'),
                                               dCt_adj = ifelse(filtered2$Is_MspI, NA, filtered2$dCt_adj.limma_ctrl_p)))

```

## Amplicon analysis
### UMAP plot by amplicon genotyping
```{r, dev=c('png', 'pdf'), fig.height=5, fig.width=10}

DimPlot(filtered2, reduction = "umap",
        group.by = 'genotype_srsf2_2',
        split.by = 'group', cols = col_genotype) +
  plot_layout(ncol = 1, nrow = 1,
              widths = unit(3*2, "in"),
              heights = unit(3, "in") )

DimPlot(filtered2, reduction = "umap",
        group.by = 'genotype_tet2a_2',
        split.by = 'group', cols = col_genotype) +
  plot_layout(ncol = 1, nrow = 1,
              widths = unit(3*2, "in"),
              heights = unit(3, "in") )

DimPlot(filtered2, reduction = "umap", 
        group.by = 'genotype_tet2b_2', 
        split.by = 'group', cols = col_genotype) +
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3*2, "in"),   
              heights = unit(3, "in") )

```

### Average SRSF2 mut ratio 
```{r, dev=c('png', 'pdf'), fig.width=15, fig.height=30}

geno_levels  <- c("wt/wt", "mut/wt", "mut/mut", "unidentifiable")
treat_levels <- c("C4D1", "C4D8") 

ord <- c(0,1,2,3,4) # reorder seurat cluster 

source('../script/plot_genotype_bar.R')

p <- plot_genotype_bar(filtered2@meta.data, 'srsf2')

patchwork::wrap_plots(p, ncol = 1)
```


### Average TET2A mut ratio 

```{r, dev=c('png', 'pdf'), fig.width=15, fig.height=30}
p <- plot_genotype_bar(filtered2@meta.data, 'tet2a')
patchwork::wrap_plots(p, ncol = 1)
```

### Average TET2B mut ratio 

```{r, dev=c('png', 'pdf'), fig.width=15, fig.height=30}
p <- plot_genotype_bar(filtered2@meta.data, 'tet2b')
patchwork::wrap_plots(p, ncol = 1)
```


### Concordance of FACS and Azimuth clustering

We can now manually assigne most seurat clusters.  

```
1/3: C4D1
0/2: C4D8
4: Mixed likely lymphoid
```
```
0: MPP1/2
1: HSC/MPP1/2
2: MEP/Ery
3: MPP1
4: LMPP/MPP1
```

```{r, dev=c('png', 'pdf'), fig.width=15, fig.height=20}

print(ord)

filtered2@meta.data <- filtered2@meta.data %>%
  mutate(seurat_clusters = factor(seurat_clusters, levels = ord))

sankey_df <- data.frame(
  # singleR.Blueprint = filtered2$SingleR.Blueprint.fine,
  # singleR.Novershtern = filtered2$SingleR.Novershtern.main,
  cluster = factor(filtered2$seurat_clusters, levels = ord),
  devcell = factor(filtered2$predicted.RNA_clusters, levels = names(col_ref)), ## change to Azimuth annotation
  FACS = factor(filtered2$celltype_merge, levels = names(col_facs)),
  rowname = colnames(filtered2), 
  source = filtered2$Cell_Type, 
  treatment = filtered2$Treatment,
  group = factor(filtered2$group, levels = names(col_group)),
  dCt_adj = filtered2$dCt_adj,
  genotype_srsf2 = filtered2$genotype_srsf2_2,
  genotype_tet2a = filtered2$genotype_tet2a_2,
  genotype_tet2b = filtered2$genotype_tet2b_2,
  Is_MspI = filtered2$Is_MspI
  )

# all-in-one comparison

sankey_df %>%
  # filter(source != 'MDSL') %>%
  group_by(source, treatment) %>%
  group_modify(~make_long(.x, group, cluster, devcell)) %>%
  mutate(node = factor(node, 
                       levels = c(rev(ord), 
                                  rev(names(col_ref)),
                                  names(col_group))) 
  ) %>%
  ggplot(aes(x = x, 
             next_x = next_x, 
             node = node, 
             next_node = next_node,
             fill = node,
             label = node)) +
  geom_sankey(flow.alpha = .6,
              node.color = "gray30") +
  geom_sankey_label(size = 3, color = "white", fill = "gray40") +
  # scale_fill_viridis_d(drop = FALSE) +
  scale_fill_manual(values = c(col_cluster, 
                               col_ref,
                               col_group))   

## compare dev cell vs transcriptome based clusters
sankey_df %>%
  filter(source != 'MDSL') %>%
  group_by(source, treatment) %>%
  group_modify(~make_long(.x, cluster, devcell)) %>%
  mutate(node = factor(node, 
                       levels = c(rev(ord), 
                                  rev(names(col_ref)))) 
  ) %>% 
  ggplot(aes(x = x, 
             next_x = next_x, 
             node = node, 
             next_node = next_node,
             fill = node,
             label = node)) +
  geom_sankey(flow.alpha = .6,
              node.color = "gray30") +
  geom_sankey_label(size = 3, color = "white", fill = "gray40") +
  # scale_fill_viridis_d(drop = FALSE) +
  scale_fill_manual(values = c(col_cluster, 
                               col_ref
  ))   

## compare dev cell vs facs clusters
sankey_df %>%
  filter(source != 'MDSL') %>%
  group_by(source, treatment) %>%
  group_modify(~make_long(.x, FACS, devcell)) %>%
  mutate(node = factor(node, 
                       levels = c(
                         rev(names(col_ref)),
                         rev(names(col_facs))) %>% rev() %>% unique() %>% rev() ) 
  ) %>% 
  ggplot(aes(x = x, 
             next_x = next_x, 
             node = node, 
             next_node = next_node,
             fill = node,
             label = node)) +
  geom_sankey(flow.alpha = .6,
              node.color = "gray30") +
  geom_sankey_label(size = 3, color = "white", fill = "gray40") +
  # scale_fill_viridis_d(drop = FALSE) +
  scale_fill_manual(values = c(col_facs, 
                               col_ref
  ))  

## compare facs clusters to seurat clusters
sankey_df %>%
  filter(source != 'MDSL') %>%
  group_by(source, treatment) %>%
  group_modify(~make_long(.x, FACS, cluster)) %>%
  mutate(node = factor(node, 
                       levels = c(
                         rev(names(col_cluster)),
                         rev(names(col_facs))) %>% unique() ) 
  ) %>% 
  ggplot(aes(x = x, 
             next_x = next_x, 
             node = node, 
             next_node = next_node,
             fill = node,
             label = node)) +
  geom_sankey(flow.alpha = .6,
              node.color = "gray30") +
  geom_sankey_label(size = 3, color = "white", fill = "gray40") +
  # scale_fill_viridis_d(drop = FALSE) +
  scale_fill_manual(values = c(col_facs, 
                               col_cluster
  ))  

```

### Sankey plot for each group (cell_type + treatment)
```{r, dev=c('png', 'pdf'), fig.height=10, fig.width=20}

for (i in unique(sankey_df$group)[1:2]) {
  
  combined_levels <- c(names(col_celltype),
                       ord)
  
  ## alluvial plot
  
  plot_df <- sankey_df %>%
    filter(group == i & FACS != "other" & !is.na(FACS)) %>%
    mutate(
      FACS    = paste0("FACS_", as.character(FACS)),
      # cluster = paste0("CL_", as.character(cluster)),
      devcell = paste0("DEV_", as.character(devcell))
    ) %>%
    dplyr::select(rowname, FACS, cluster, devcell, dCt_adj) %>%
        mutate(devcell = factor(devcell, levels = combined_levels),
           cluster = factor(cluster, levels = combined_levels),
           FACS = factor(FACS, levels = combined_levels)) 
  
  y_max <- nrow(plot_df)
  
  p_alluvial <- plot_df %>%
    # dplyr::select(2:4) %>%
    dplyr::count(FACS, cluster, devcell) %>%
    to_lodes_form(axes=1:3, id="id") %>%
    ggplot(aes(x=x, y=n, stratum=stratum, alluvium=id,
               fill=stratum, label=stratum)) +
    geom_flow(stat = "alluvium" #, color="darkgray"
    )+
    geom_stratum(alpha=0.8) +
    geom_text(
      aes(label = ifelse(x == 'cluster', as.character(stratum), NA)),
      stat = "stratum") +
    ggrepel::geom_text_repel(
      aes(label = ifelse(x == 'FACS', as.character(stratum), NA)),
      stat = "stratum", size = 3, direction = "y", nudge_x = -.8
    ) +
    ggrepel::geom_text_repel(
      aes(label = ifelse(x == 'devcell', as.character(stratum), NA)),
      stat = "stratum", size = 3, direction = "y", nudge_x = .8
    ) +
    theme(legend.position = "left") + 
    scale_fill_manual(values = c(col_celltype, col_cluster)) +
    scale_y_continuous(limits = c(0, y_max), expand = expansion(mult = c(0, 0))) +
    coord_cartesian(clip = "off") +
    ggtitle(paste(i, '\nImmunophenotye vs Seurat cluster vs Dev Cell annotation Alluvial plot'))
  
  p_waterfall <- plot_df %>%
  arrange(devcell, cluster, FACS, desc(dCt_adj)) %>%                 # keep your ordering
  mutate(rowname = fct_rev(factor(rowname, levels = rowname))) %>%      # freeze order
  ggplot(aes(y = rowname)) +
  # bar length encodes value; start at lower limit so lengths are comparable
  geom_segment(
    aes(x = 9.5, xend = dCt_adj, yend = rowname, colour = dCt_adj),
    linewidth = 2, lineend = "round", na.rm = TRUE
  ) +
  geom_point(aes(x = dCt_adj, colour = dCt_adj), size = 1.4, na.rm = TRUE) +
  # scale_x_continuous(limits = c(10.5, 16), expand = expansion(mult = c(0, 0.02))) +
  scale_y_discrete(expand = expansion(add = c(0, 0))) +
  coord_cartesian(xlim = c(3,12), clip = "off") + # show full range
  scale_colour_viridis_c(
    direction = -1, limits = c(3,12),
    oob = squish, na.value = "grey90"
  ) +
  labs(x = "dCt_adj", y = NULL, colour = "dCt_adj") +
  theme_minimal() +
  theme(
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  ggtitle(paste(i, "\ndCt Waterfall"))
  
  combined_plot <- p_alluvial + p_waterfall +
    plot_layout(ncol = 2,
                 widths = c(3, 0.5))
  
  # use explicitly print to render in Quarto
  print(combined_plot)
  
}

```


### Sankey plot by SRSF2 genotypes

```{r,  dev=c('png', 'pdf'), fig.height=10, fig.width=20}

plot_df <- sankey_df %>%
  filter(FACS != "other" & !is.na(FACS)) %>%
  dplyr::select(group, rowname, FACS, cluster, devcell, dCt_adj, 
                genotype_srsf2, genotype_tet2a, genotype_tet2b) %>%
  mutate(#devcell = factor(devcell, levels = combined_levels),
         cluster = factor(cluster, levels = ord),
         FACS = factor(FACS, levels = names(col_facs))) 

plot_genotype_alluvial <- function(df, title_suffix = "") {
  # stopifnot(is_alluvial(df, axes = 1:2, silent = TRUE) || TRUE) # harmless
  
  df %>%
    count(FACS, cluster, name = "n") %>%                 # one genotype subset already
    to_lodes_form(axes = 1:2, id = "id") %>%
    ggplot(aes(x = x, y = n, stratum = stratum, alluvium = id,
               fill = stratum, label = stratum)) +
    geom_flow(stat = "alluvium") +
    geom_stratum(alpha = 0.8) +
    ggrepel::geom_text_repel(
      aes(label = ifelse(x == "FACS", as.character(stratum), NA)),
      stat = "stratum", size = 3, direction = "y", nudge_x = -0.8
    ) +
    ggrepel::geom_text_repel(
      aes(label = ifelse(x == "cluster", as.character(stratum), NA)),
      stat = "stratum", size = 3, direction = "y", nudge_x =  0.8
    ) +
    scale_fill_manual(values = c(col_facs, col_cluster)) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    ggtitle(title_suffix) +
    theme(legend.position = "right", 
          axis.title.x = element_blank() )
}

# for C4D1
plots <- plot_df %>%
  filter(genotype_srsf2 != "unidentifiable",
         group == 'C4D1_C4D1' ) %>%
  group_split(genotype_srsf2) %>%
  lapply(function(dd) {
    gname <- unique(dd$genotype_srsf2)
    plot_genotype_alluvial(dd, gname)
  })

# Show as a row (or use wrap_plots(plots, ncol = 2))
p_alluvial <- patchwork::wrap_plots(plots, ncol = length(plots)) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "C4D1 Immunophenotype vs Seurat cluster (by SRSF2 genotype)") &                # collect a single legend
  theme(legend.position = "right")   
p_alluvial


## for C4D8
plots <- plot_df %>%
  filter(genotype_srsf2 != "unidentifiable",
         group == 'C4D8_C4D8' ) %>%
  group_split(genotype_srsf2) %>%
  lapply(function(dd) {
    gname <- unique(dd$genotype_srsf2)
    plot_genotype_alluvial(dd, gname)
  })

# Show as a row (or use wrap_plots(plots, ncol = 2))
p_alluvial <- patchwork::wrap_plots(plots, ncol = length(plots)) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "C4D8 Immunophenotype vs Seurat cluster (by SRSF2 genotype)") &                # collect a single legend
  theme(legend.position = "right")   
p_alluvial


```

## LINE1 qPCR results plot by treatment and cell types

```{r, dev=c('png', 'pdf'), fig.width=15, fig.height=10}
lims <- c(3,12)

## 1454 LINE1 pass cells
filtered2_pass <- subset(filtered2, subset = keep_median & dCt > 0)

patchwork::wrap_plots( FeaturePlot( filtered2_pass,
                                    features=c("dCt_adj"),
                                    combine=FALSE)) +
  plot_layout(ncol = 1, nrow = 1,
              widths = unit(c(3,3), "in"),
              heights = unit(3, "in") ) &
  theme(legend.position = "right") &
  scale_color_viridis_c(direction = -1,
                        limits = lims,
    oob = squish, # squish values outside limits
    na.value = "grey90") &
  ggtitle('UMAP of LINE1 and RNAseq passed cells')

patchwork::wrap_plots( FeaturePlot( filtered2_pass,
                                    features=c("dCt_adj"),
                                    split.by = 'group',
                                    combine=FALSE)) +
  plot_layout(ncol = 3, nrow = 2,
              widths = unit(c(3,3), "in"),
              heights = unit(3, "in") ) &
  theme(legend.position = "right") &
    scale_color_viridis_c(direction = -1,
                        limits = lims,
    oob = squish, # squish values outside limits
    na.value = "grey90")


VlnPlot(filtered2_pass,
        features = 'dCt_adj',
        group.by = 'Cell_Type', split.by = 'Treatment',
        cols = col_treatment,
        split.plot = T) +
  scale_y_continuous(limits = lims)

VlnPlot(filtered2_pass,
        features = 'dCt_adj',
        group.by = 'celltype_merge',
        cols = col_facs) +
  scale_y_continuous(limits = lims)

VlnPlot(filtered2_pass,
        features = 'dCt_adj',
        group.by = 'predicted.RNA_clusters',
        cols = col_ref) +
  scale_y_continuous(limits = lims)

VlnPlot(filtered2_pass,
        features = 'dCt_adj',
        group.by = 'seurat_clusters',
        cols = col_cluster) +
  scale_y_continuous(limits = lims)

```


### LINE1 by FACS subtype
```{r, dev=c('png', 'pdf')}
# All data (LINE1 and RNAseq passed)

source('../script/vlnplot_padj.R')

df <- FetchData(filtered2_pass,
                vars = c('dCt_adj','celltype_merge','Treatment'))

df_filtered <- df %>%
  filter(!is.na(celltype_merge) & celltype_merge != 'other') %>%
  filter(!is.na(dCt_adj)) %>%           # <--- CRITICAL STEP: Remove MspI well with NA value now
  group_by(celltype_merge, Treatment) %>%
  filter(n() >= 2) %>%
  ungroup()

compare_means(
  formula         = dCt_adj ~ Treatment,
  group.by        = "celltype_merge",
  data            = df_filtered,
  method          = 't.test',
  p.adjust.method = "fdr"
)

vlnplot_padj(filtered2_pass,
             c('dCt_adj','celltype_merge','Treatment','genotype_srsf2_2'),
             facetorder = names(col_facs),
             fill = col_group,
             lims = lims)
vlnplot_padj(filtered2_pass,
             c('dCt_adj','celltype_merge','Treatment','genotype_tet2a_2'),
             facetorder = names(col_facs),
             fill = col_group, 
             lims = lims)
vlnplot_padj(filtered2_pass,
             c('dCt_adj','celltype_merge','Treatment','genotype_tet2b_2'),
             facetorder = names(col_facs),
             fill = col_group,
             lims = lims)

```

### LINE1 by Seurat cluster

```{r, dev=c('png', 'pdf')}
# All data (LINE1 and RNAseq passed)

df <- FetchData(filtered2_pass,
                vars = c('dCt_adj','seurat_clusters','Treatment'))

df_filtered <- df %>%
  # filter(!is.na(celltype_merge) & celltype_merge != 'other') %>%
  filter(!is.na(dCt_adj)) %>%           # <--- CRITICAL STEP: Remove MspI well with NA value now
  group_by(seurat_clusters, Treatment) %>%
  filter(n() >= 2) %>%
  ungroup()

compare_means(
  formula         = dCt_adj ~ Treatment,
  group.by        = "seurat_clusters",
  data            = df_filtered,
  method          = 't.test',
  p.adjust.method = "fdr"
)

vlnplot_padj(filtered2_pass,
             c('dCt_adj','seurat_clusters','Treatment','genotype_srsf2_2'),
             facetorder = ord,  # use John suggested order
             fill = col_group,
             lims = lims)
vlnplot_padj(filtered2_pass,
             c('dCt_adj','seurat_clusters','Treatment','genotype_tet2a_2'),
             facetorder = ord,
             fill = col_group,
             lims = lims)
vlnplot_padj(filtered2_pass,
             c('dCt_adj','seurat_clusters','Treatment','genotype_tet2b_2'),
             facetorder = ord,
             fill = col_group,
             lims = lims)

```

### LINE1 by Dev Cell annotation (Azimuth)

```{r, dev=c('png', 'pdf')}
# All data (LINE1 and RNAseq passed)

df <- FetchData(filtered2_pass,
                vars = c('dCt_adj','predicted.RNA_clusters','Treatment'))

df_filtered <- df %>%
  # filter(!is.na(celltype_merge) & celltype_merge != 'other') %>%
  filter(!is.na(dCt_adj)) %>%           # <--- CRITICAL STEP: Remove MspI well with NA value now
  group_by(predicted.RNA_clusters, Treatment) %>%
  filter(n() >= 2) %>%
  ungroup()

compare_means(
  formula         = dCt_adj ~ Treatment,
  group.by        = "predicted.RNA_clusters",
  data            = df_filtered,
  method          = 't.test',
  p.adjust.method = "fdr"
)

vlnplot_padj(filtered2_pass,
             c('dCt_adj','predicted.RNA_clusters','Treatment','genotype_srsf2_2'),
             facetorder = names(col_ref),
             fill = col_group,
             lims = lims)
vlnplot_padj(filtered2_pass,
             c('dCt_adj','predicted.RNA_clusters','Treatment','genotype_tet2a_2'),
             facetorder = names(col_ref),
             fill = col_group,
             lims = lims)
vlnplot_padj(filtered2_pass,
             c('dCt_adj','predicted.RNA_clusters','Treatment','genotype_tet2b_2'),
             facetorder = names(col_ref),
             fill = col_group,
             lims = lims)

```

### LINE1 by cluster and FACS 
```{r, fig.width=12, fig.height=12, dev=c('png', 'pdf')}

# 2) Apply levels and plot
filtered2_pass@meta.data %>%
  filter(!Is_MspI) %>%
  ggplot(aes(x = seurat_clusters, y = dCt_adj, fill = seurat_clusters)) +
  geom_violin(trim = FALSE, scale = "width",
              alpha = 0.9, width = 0.9, colour = NA, show.legend = FALSE) +
  geom_boxplot(width = 0.15, outlier.alpha = 0.25, colour = "grey50",
               fill = "white", alpha = 0.9, show.legend = FALSE) +
  facet_wrap(~ celltype_merge) +
  scale_fill_manual(
    values   = col_cluster,
    na.value = na_grey,
    name     = 'Seurat Cluster'
  ) +
  labs(x = "Cluster (ordered by cluster mean dCt_adj)", y = "dCt_adj") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  force_panelsizes(rows = unit(3, 'inch'),
                   cols = unit(3, 'inch'))


```

### LINE1 by cluster and FACS (SRSF2 genotype)
```{r, fig.width=8, fig.height=18, dev=c('png', 'pdf')}
filtered2_pass@meta.data %>%
  filter(!Is_MspI) %>%
  filter(genotype_srsf2 != 'unidentifiable') %>%
  ggplot(aes(x = seurat_clusters, y = dCt_adj, fill = genotype_srsf2)) +
  # paired violins per cluster
  geom_violin(
    aes(group = interaction(seurat_clusters, genotype_srsf2)),
    trim = FALSE, scale = "width",
    width = 0.9, colour = NA,
    position = position_dodge(width = 0.8), alpha = 0.8
  ) +
  # paired boxplots exactly overlaid on the violins
  geom_boxplot(
    aes(group = interaction(seurat_clusters, genotype_srsf2)),
    width = 0.15, outlier.alpha = 0.3, colour = "grey40",
    fill = "white", alpha = 0.9,
    position = position_dodge(width = 0.8)
  ) +
  facet_grid(celltype_merge ~ Treatment) +
  scale_fill_manual(
    values   = col_genotype,
    na.value = na_grey,
    name     = 'Genotype'
  ) +
  labs(
    x = "Cluster",
    y = "dCt_adj"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  force_panelsizes(rows = unit(3, 'inch'),
                   cols = unit(3, 'inch'))

```


```{r}
sessionInfo()
```
