---
title: "SCIMETAR single cell analysis (Flitered by RPKM) add LINE1, FACS and amplicon data"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
    code-overflow: wrap
    number-sections: true
    number-depth: 3
execute: 
  cache: true
  warning: false
  error: false
---

## Load filtered and annotated seurat object 

```{r, dev=c('png', 'pdf')}
library(matrixStats)
library(ggfortify)
library(Seurat)
library(edgeR)
library(tidyverse)
library(DT)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggpubr)
library(EnhancedVolcano)
library(ggh4x)
library(viridis)
library(patchwork)
library(readxl)
library(RColorBrewer)
library(UpSetR)
library(data.table)
library(ggsankey)
library(ggalluvial)
library(ggrepel)
library(scales)
library(ggsci)
```


```{r}

filtered2 <- readRDS("output/filtered2.rds")

na_grey   <- "grey85"            # any NA in celltype_merge

# use the Dev Cell colors extracted, same order as the weblink
col_ref <- c(
  HSC   = "#ED6D52",
  MPP1  = "#F5C3CB",
  MPP2  = "#DF87E8",
  LMPP  = "#5DCBCF",
  MLP   = "#4A895B",
  MEP   = "#8E71D5",
  Neu1  = "#F2A93B",
  B_NK1 = "#75FB8C",
  Neu2  = "#C3884D",
  MD    = "#EFE697",
  EBM   = "#FADFB3",
  Mk    = "#4A68DA",
  Ery   = "#97CCE7",
  B_NK2 = "#A6EB99"
)

col_celltype <- c(
  DEV_HSC   = "#ED6D52", FACS_HSC = "#ED6D52", 
  DEV_MPP1  = "#F5C3CB", #FACS_MHSC = "#DF87E8", 
  FACS_MPP = "#F5C3CB",
  DEV_MPP2  = "#DF87E8",
  DEV_MEP   = "#8E71D5", FACS_MEP = "#8E71D5",
  DEV_Ery   = "#97CCE7",
  DEV_Mk    = "#4A68DA",
  DEV_LMPP  = "#5DCBCF", FACS_LMPP = "#5DCBCF", FACS_CMP = "#F2A93B", 
  DEV_MLP   = "#4A895B", FACS_CLP = "#4A895B",
  DEV_B_NK1 = "#75FB8C",
  DEV_B_NK2 = "#A6EB99",
  DEV_Neu1  = "#F2A93B", FACS_GMP = "#C3884D", 
  DEV_Neu2  = "#C3884D",
  DEV_MD    = "#EFE697",
  DEV_EBM   = "#FADFB3"
)

col_celltype <- col_celltype[c(
  "FACS_HSC",'DEV_HSC',
  "FACS_MHSC","FACS_MPP","DEV_MPP1","DEV_MPP2",
  "FACS_CMP","FACS_LMPP",'DEV_LMPP',
  "FACS_GMP", "DEV_Neu1","FACS_MEP",'DEV_MEP',"FACS_CLP","DEV_MLP",
  "DEV_Neu2","DEV_B_NK1",
  "DEV_MD","DEV_EBM","DEV_Ery","DEV_Mk")]

# col_facs <- setNames(pal_npg("nrc")(10)[1:9], filtered2@meta.data$celltype_merge %>% factor %>% levels)
# col_facs["other"] <- "grey50"        # explicit grey for “other”
col_facs <- c(
  HSC = "#ED6D52",
  #MHSC = "#DF87E8", 
  MPP = "#F5C3CB",
  LMPP = "#5DCBCF",
  CMP = "#F2A93B", 
  CLP = "#4A895B",
  MEP = "#8E71D5",
  GMP = "#C3884D",
  other = "grey50"
)

# col_group <- setNames(brewer.pal(12, "Paired")[c(3:6)], filtered2@meta.data$group %>% factor %>% levels)
col_group <- c("C4D1_C4D1" = "#80cdc1", 
               "C4D8_C4D8" = "#018571"
               # "C4D8_C4D8" = "#dfc27d",
               # 'Primary_Patient_AZA' = "#a6611a"
               )

col_cluster <- setNames(brewer.pal(12, "Set3")[c(1,3:8,10,11)], 0:8)

col_treatment <- setNames(brewer.pal(2, "Set2")[1:2], c('C4D1','C4D8'))

col_source <- setNames(c("#018571" , "#a6611a"), c('GR', 'Primary_Patient'))

col_phase <- c(G1='#663399', G2M='#E66100', S='#2A934E')

col_genotype <- c(
  "unidentifiable" = '#d9d9d9',
  "wt/wt" = '#253494',
  "mut/wt" = '#dd1c77',
  # "mut/mut" = '#980043',
  "mut/mut" = '#dd1c77'
)

# 2 group genotype
filtered2 <- AddMetaData(filtered2, 
                         metadata = data.frame(genotype_srsf2_2 = str_replace(filtered2$genotype_srsf2, 'mut/mut','mut/wt'),
                                               genotype_tet2a_2 = str_replace(filtered2$genotype_tet2a, 'mut/mut','mut/wt'),
                                               genotype_tet2b_2 = str_replace(filtered2$genotype_tet2b, 'mut/mut','mut/wt'),
                                               dCt_adj.limma_ctrl_p = ifelse(filtered2$Is_MspI, NA, filtered2$dCt_adj.limma_ctrl_p)))

## 1454 LINE1 pass cells, 
filtered2_pass <- subset(filtered2, subset = keep_median & dCt > 0)

```

## DEG analysis

### DEG analysis in each cluster comparing the top 1/3 vs bottom 1/3 based on LINE1
* Only remove failed RNAseq cells.  
* Cell.1 is top and Cell.2 is bottom.  
* Positive logFC means higher expression in top.
* Use limma_ctrl_p 

```{r,fig.width=12, fig.height=7,  dev=c('png', 'pdf')}
# 1345 cells removing MspI
filtered2_pass <- subset(filtered2_pass, subset = Is_MspI == FALSE)

gencode_gtf <- readRDS('/home/users/allstaff/yan.a/lab_davidson/yan.a/ref/gencode/gencodev44.rds')

genes <- gencode_gtf %>% plyranges::filter(type=='gene') %>%
  plyranges::select(gene_id, gene_name) %>% data.frame()

m <- genes$gene_id %>%
  str_remove("\\..*") %>%
  unique() %>%
  bitr(fromType = "ENSEMBL",
       toType   = c("ENTREZID","SYMBOL"),
       OrgDb    = org.Hs.eg.db)

m1 <- m %>%
  left_join(genes %>% transmute(ENSEMBL = str_remove(gene_id, "\\..*"),
                                gene_name),
            by = "ENSEMBL") %>%
  mutate(match_name = (!is.na(gene_name) & SYMBOL == gene_name)) %>%
  arrange(ENSEMBL, desc(match_name), is.na(ENTREZID)) %>%  # prefer name match, then non-NA
  distinct(ENSEMBL, .keep_all = TRUE) %>%
  dplyr::select(-gene_name, -match_name)

genes <- genes %>% 
  mutate(ENSEMBL = str_remove(gene_id, "\\..*")) %>% 
  left_join(m1, by = 'ENSEMBL')

deglist.cl2 <- lapply(0:4, function(x){
  sub <- subset(filtered2_pass, subset = seurat_clusters == x)
  
  median <- median(sub$dCt_adj.limma_ctrl_p)
  tercicle <- quantile(sub$dCt_adj.limma_ctrl_p, probs = seq(0, 1, length.out = 4))
  
  sub <- AddMetaData(sub, 
                     metadata = data.frame(Median = ifelse(sub$dCt_adj.limma_ctrl_p >= median, 'top','bottom'),
                                           Terciles = ifelse(sub$dCt_adj.limma_ctrl_p >= tercicle[3], 'top', 
                                                             ifelse(sub$dCt_adj.limma_ctrl_p < tercicle[2], 'bottom', 'other'))))
  deg <- tryCatch({
    FindMarkers(
      object        = sub,
      ident.1       = "top",               # which treatment to test
      ident.2       = "bottom",              # what to compare against, baseline
      group.by      = "Terciles",         # which metadata column defines AZA vs DMSO
      # subset.ident  = "Primary_CB34",   # first subset to only these cells
      only.pos      = FALSE
    ) %>% rownames_to_column(var = 'gene_name') %>%
      left_join(genes, by = 'gene_name')
  }, error = function(e) data.frame())
})
 
names(deglist.cl2) <- paste0('Cluster', 0:4)

gene.list.cl2 <- lapply(deglist.cl2, function(x) {
  if (nrow(x) != 0) {
      x %>% filter(p_val < 0.05) %>% pull(ENTREZID)
  }
})

fromList(gene.list.cl2) %>%
  UpSetR::upset()

go.cl2 <- gene.list.cl2 %>%
  compareCluster(fun = 'enrichGO', OrgDb = org.Hs.eg.db, ont = 'BP', readable = T) %>%
  clusterProfiler::simplify()

dotplot(go.cl2)

# # collapse all DE tables into one and summarise a single logFC per gene
# fc <- bind_rows(deglist.cl2, .id = "Cluster") %>%
#   filter(!is.na(SYMBOL), SYMBOL != "") %>%
#   group_by(SYMBOL) %>%
#   summarise(logFC = mean(avg_log2FC, na.rm = TRUE), .groups = "drop") %>%
#   deframe()   # named numeric vector: names are SYMBOL

cnetplot(go.cl2, 
         #color.params = list(foldChange = fc),  # preferred (newer enrichplot)
         #fc_threshold = 0
         )

lapply(names(deglist.cl2), function(x) {
  if (nrow(deglist.cl2[[x]]) != 0) {
      deglist.cl2[[x]] %>% mutate(comparison = x) %>%
    filter(p_val < 0.05)
  }
}) %>% rbindlist() %>%
  write.csv(., file = 'output/line1_deg_withincluster2.csv', quote = F, row.names = F)

write.csv(go.cl2@compareClusterResult, 
           file = 'output/line1_deg_withincluster_go2.csv', quote = F, row.names = F)
```


### DEG analysis in each timepoint comparing the top 1/3 vs bottom 1/3 based on LINE1
* Only remove failed RNAseq cells.  
* Cell.1 is top and Cell.2 is bottom.  
* Positive logFC means higher expression in top.
* Use limma_ctrl_p 

```{r,fig.width=12, fig.height=7,  dev=c('png', 'pdf')}
# 1345 cells removing MspI

deglist.tp2 <- lapply( c('C4D1', 'C4D8', 'Both'), function(x){
    if (x != 'Both') {
    sub <- subset(filtered2_pass, subset = Cell_Type == x)
  } else {
    sub <- filtered2_pass
  }
  
  median <- median(sub$dCt_adj.limma_ctrl_p)
  tercicle <- quantile(sub$dCt_adj.limma_ctrl_p, probs = seq(0, 1, length.out = 4))
  
  sub <- AddMetaData(sub, 
                     metadata = data.frame(Median = ifelse(sub$dCt_adj.limma_ctrl_p >= median, 'top','bottom'),
                                           Terciles = ifelse(sub$dCt_adj.limma_ctrl_p >= tercicle[3], 'top', 
                                                             ifelse(sub$dCt_adj.limma_ctrl_p < tercicle[2], 'bottom', 'other'))))
  deg <- tryCatch({
    FindMarkers(
      object        = sub,
      ident.1       = "top",               # which treatment to test
      ident.2       = "bottom",              # what to compare against, baseline
      group.by      = "Terciles",         # which metadata column defines AZA vs DMSO
      # subset.ident  = "Primary_CB34",   # first subset to only these cells
      only.pos      = FALSE
    ) %>% rownames_to_column(var = 'gene_name') %>%
      left_join(genes, by = 'gene_name')
  }, error = function(e) data.frame())
})
 
names(deglist.tp2) <- c('C4D1', 'C4D8', 'Both')

gene.list.tp2 <- lapply(deglist.tp2, function(x) {
  if (nrow(x) != 0) {
      x %>% filter(p_val < 0.05) %>% pull(ENTREZID)
  }
})

fromList(gene.list.tp2) %>%
  UpSetR::upset()

go.tp2 <- gene.list.tp2 %>%
  compareCluster(fun = 'enrichGO', OrgDb = org.Hs.eg.db, ont = 'BP', readable = T) %>%
  clusterProfiler::simplify()

dotplot(go.tp2)

cnetplot(go.tp2, 
         #color.params = list(foldChange = fc),  # preferred (newer enrichplot)
         #fc_threshold = 0
         )

lapply(names(deglist.tp2), function(x) {
  if (nrow(deglist.tp2[[x]]) != 0) {
      deglist.tp2[[x]] %>% mutate(comparison = x) %>%
    filter(p_val < 0.05)
  }
}) %>% rbindlist() %>%
  write.csv(., file = 'output/line1_deg_timepoint2.csv', quote = F, row.names = F)

write.csv(go.tp2@compareClusterResult, 
           file = 'output/line1_deg_timepoint_go2.csv', quote = F, row.names = F)

gene.list.tp2.split <- c(lapply(deglist.tp2, function(x) {
  if (nrow(x) != 0) { x %>% filter(p_val < 0.05, avg_log2FC > 0) %>% pull(ENTREZID) }
}),  lapply(deglist.tp2, function(x) {
  if (nrow(x) != 0) { x %>% filter(p_val < 0.05, avg_log2FC < 0) %>% pull(ENTREZID) }
}))
names(gene.list.tp2.split) <- paste0(names(gene.list.tp2.split), rep(c('_up', '_down'), each = 3))

fromList(gene.list.tp2.split) %>%
  UpSetR::upset(sets = names(gene.list.tp2.split))

go.tp2.split <- gene.list.tp2.split %>%
  compareCluster(fun = 'enrichGO', OrgDb = org.Hs.eg.db, ont = 'BP', readable = T) %>%
  clusterProfiler::simplify()

dotplot(go.tp2.split)
cnetplot(go.tp2.split
         # , foldChange = setNames(deg.pp$avg_log2FC, deg.pp$ENTREZID)
         )

write.csv(go.tp2.split@compareClusterResult, 
           file = 'output/line1_deg_timepoint_go_splitdirection2.csv', quote = F, row.names = F)
```

