---
title: "SCIMETAR single cell analysis (Flitered by RPKM) add LINE1, FACS and amplicon data"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
    code-overflow: wrap
    number-sections: true
    number-depth: 3
execute: 
  cache: true
  warning: false
  error: false
---

## Load filtered cells 


```{r, dev=c('png', 'pdf')}
library(matrixStats)
library(ggfortify)
library(Seurat)
library(edgeR)
library(tidyverse)
library(DT)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggpubr)
library(EnhancedVolcano)
library(ggh4x)
library(viridis)
library(patchwork)
library(readxl)
library(RColorBrewer)
library(UpSetR)
library(data.table)
library(ggsankey)
library(ggalluvial)
library(ggrepel)
library(scales)
library(ggsci)
library(Azimuth)
```

```{r}
seuratobj <- readRDS('output/seurat_zumi_rpkm_unfilt2.rds')
```


I use nFeature_RNA > 2000 & not empty & remove DEAD cells in FACS annotation.  
Also remove MDSL cells. Use new FACS annotation without IMM subtype.  
There are 1734 cells passed the filter.  
```{r}
## load FACS data
All_cell_assignments <- read_excel("data/All_cell_assignments_allCD34pos.xlsx") %>%
  separate(Plate_Well, c('Plate', NA,NA,'Well_ID'), remove = F) %>%
  mutate(Plate = str_remove(Plate, '^P') %>% as.numeric())

## load LINE1 data
pcr <- read.csv('output/LINE1_qPCR_corrected.csv', row.names = 1) 

## load amplicond data, new version
amplicon <- read.csv('output/genotyped_data_2025-12-02.csv') %>%
  dplyr::select(c(1:2,11:19)) %>%
  mutate(Plate2 = Plate, Well_ID = Well) %>%
  mutate(Plate = str_remove(Plate2, '^P') %>% as.numeric()) %>%
  mutate(across(matches("wt_|mut_"), ~ log1p(.x), 
                .names = "{.col}_log1p"))

master <- left_join(pcr, All_cell_assignments,
                    by = c('Plate','Well_ID')) %>%
  left_join(amplicon, by = c('Plate','Well_ID')) %>%
  mutate( # use healthy cell type
    celltype_merge = factor(celltype_Healthy,
                                 levels = c("HSC","MPP","LMPP","CMP","CLP","MEP","GMP","other",'DEAD')),
         group = factor(group, 
                        levels = c("Primary_CB34_DMSO","Primary_CB34_AZA","Primary_Patient_DMSO","Primary_Patient_AZA" )))

rownames(master) <- master$Transcriptome.ID
write.csv(master, 'output/master.csv', quote = F, row.names = F)

seuratobj <- AddMetaData(seuratobj, 
                         master)

filtered2 <- subset(seuratobj, subset = nFeature_RNA > 2000 & Source != 'MD' &
                      empty_well == FALSE & celltype_merge != 'DEAD' & !is.na(celltype_merge))

filtered2@meta.data %>% dplyr::count(Cell_Type, Treatment)
```


```{r}

na_grey   <- "grey85"            # any NA in celltype_merge

# use the Dev Cell colors extracted, same order as the weblink
col_ref <- c(
  HSC   = "#ED6D52",
  MPP1  = "#F5C3CB",
  MPP2  = "#DF87E8",
  LMPP  = "#5DCBCF",
  MLP   = "#4A895B",
  MEP   = "#8E71D5",
  Neu1  = "#F2A93B",
  B_NK1 = "#75FB8C",
  Neu2  = "#C3884D",
  MD    = "#EFE697",
  EBM   = "#FADFB3",
  Mk    = "#4A68DA",
  Ery   = "#97CCE7",
  B_NK2 = "#A6EB99"
)

col_celltype <- c(
  DEV_HSC   = "#ED6D52", FACS_HSC = "#ED6D52", 
  DEV_MPP1  = "#F5C3CB", #FACS_MHSC = "#DF87E8", 
  FACS_MPP = "#F5C3CB",
  DEV_MPP2  = "#DF87E8",
  DEV_MEP   = "#8E71D5", FACS_MEP = "#8E71D5",
  DEV_Ery   = "#97CCE7",
  DEV_Mk    = "#4A68DA",
  DEV_LMPP  = "#5DCBCF", FACS_LMPP = "#5DCBCF", FACS_CMP = "#F2A93B", 
  DEV_MLP   = "#4A895B", FACS_CLP = "#4A895B",
  DEV_B_NK1 = "#75FB8C",
  DEV_B_NK2 = "#A6EB99",
  DEV_Neu1  = "#F2A93B", FACS_GMP = "#C3884D", 
  DEV_Neu2  = "#C3884D",
  DEV_MD    = "#EFE697",
  DEV_EBM   = "#FADFB3"
)

col_celltype <- col_celltype[c(
  "FACS_HSC",'DEV_HSC',
  "FACS_MHSC","FACS_MPP","DEV_MPP1","DEV_MPP2",
  "FACS_CMP","FACS_LMPP",'DEV_LMPP',
  "FACS_GMP", "DEV_Neu1","FACS_MEP",'DEV_MEP',"FACS_CLP","DEV_MLP",
  "DEV_Neu2","DEV_B_NK1",
  "DEV_MD","DEV_EBM","DEV_Ery","DEV_Mk")]

# col_facs <- setNames(pal_npg("nrc")(10)[1:9], filtered2@meta.data$celltype_merge %>% factor %>% levels)
# col_facs["other"] <- "grey50"        # explicit grey for “other”
col_facs <- c(
  HSC = "#ED6D52",
  #MHSC = "#DF87E8", 
  MPP = "#F5C3CB",
  LMPP = "#5DCBCF",
  CMP = "#F2A93B", 
  CLP = "#4A895B",
  MEP = "#8E71D5",
  GMP = "#C3884D",
  other = "grey50"
)

# col_group <- setNames(brewer.pal(12, "Paired")[c(3:6)], filtered2@meta.data$group %>% factor %>% levels)
col_group <- c("Primary_CB34_DMSO" = "#80cdc1", 
               "Primary_CB34_AZA" = "#018571",
               "Primary_Patient_DMSO" = "#dfc27d",
               'Primary_Patient_AZA' = "#a6611a"
               )

col_cluster <- setNames(brewer.pal(12, "Set3")[c(1,3:8,10,11)], 0:8)

col_treatment <- setNames(brewer.pal(2, "Set2")[1:2], c('DMSO','AZA'))

col_source <- setNames(c("#018571" , "#a6611a"), c('Primary_CB34', 'Primary_Patient'))

col_phase <- c(G1='#663399', G2M='#E66100', S='#2A934E')
```

## Top variable features and PCA/UMAP plot
Top 10 most variable genes still have HB genes.  
After clustering, there are 12 major clusters.

```{r, dev=c('png', 'pdf')}
filtered2 <- FindVariableFeatures(filtered2, 
                                  selection.method = "vst", 
                                  nfeatures = 2000)

filtered2 <- ScaleData(filtered2, features = rownames(filtered2))

filtered2 <- RunPCA(filtered2, features = VariableFeatures(object = filtered2))
VizDimLoadings(filtered2, dims = 1:2, reduction = "pca")

DimPlot(filtered2, reduction = "pca") 

filtered2 <- CellCycleScoring(object = filtered2, 
                              g2m.features = cc.genes.updated.2019$g2m.genes,
                              s.features = cc.genes.updated.2019$s.genes)

filtered2 <- FindNeighbors(filtered2, dims = 1:20)
filtered2 <- FindClusters(filtered2, resolution = 1)

filtered2 <- RunUMAP(filtered2, dims = 1:20, 
                     n.neighbors = 50)

DimPlot(filtered2, reduction = "umap", label = T, cols = col_cluster) +
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )

DimPlot(filtered2, reduction = "umap", 
        split.by = 'seurat_clusters',
        cols = col_cluster) 

DimPlot(filtered2, reduction = "umap",  group.by = 'Phase',
        cols = col_phase) + 
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )

DoHeatmap(filtered2, 
          features = c(cc.genes.updated.2019$g2m.genes, cc.genes.updated.2019$s.genes),
          group.by = 'Phase', group.colors = col_phase,
          angle = 90) 
```

### UMAP plot by plate 
to examine batch effect from plate
```{r, dev=c('png', 'pdf')}
#| fig-cap: "UMAP plot colored by different plate"
#| layout-ncol: 3

DimPlot(filtered2, reduction = "umap", group.by = 'Plate2', 
        cols = c(brewer.pal(8, 'Set1'), rep('grey90', 18)))  +
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )

DimPlot(filtered2, reduction = "umap", group.by = 'Plate2', 
        cols = c(rep('grey90', 8), brewer.pal(8, 'Set1'), rep('grey90', 10)))  +
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )

DimPlot(filtered2, reduction = "umap", group.by = 'Plate2', 
        cols = c(rep('grey90', 16), brewer.pal(4, 'Set1'), rep('grey90', 6)))  +
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )

DimPlot(filtered2, reduction = "umap", group.by = 'Plate2', 
        cols = c(rep('grey90', 20), brewer.pal(4, 'Set1'), rep('grey90', 2)))  +
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )

```


### UMAP plot by treatment and cell types
```{r, dev=c('png', 'pdf')}
#| fig-cap: "UMAP plot colored by different grouping"
#| layout-ncol: 4

DimPlot(filtered2, reduction = "umap",  group.by = 'Cell_Type',
        cols = col_source) +
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )

DimPlot(filtered2, reduction = "umap",  group.by = 'Treatment',
        cols = col_treatment) +
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )

DimPlot(filtered2, reduction = "umap",  group.by = 'group',
        cols = col_group) +
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )

DimPlot(filtered2, reduction = "umap",  group.by = 'celltype_merge',
        cols = col_facs,   na.value = na_grey) +  
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )
```


```{r, dev=c('png', 'pdf'), fig.width=7, fig.width=7}

for (group in names(col_group)) {
  
  # Create highlight column
  filtered2$highlighted_group <- ifelse(filtered2$group == group, group, NA)
  
  # Define colours
  highlight_cols <- c(col_group[group], "grey95")
  names(highlight_cols) <- c(group, "NA")
  
  # Reorder cells: grey (NA) first, highlighted cells last
  ordered_cells <- filtered2@meta.data %>%
    mutate(cell = rownames(.)) %>%
    arrange(!is.na(highlighted_group)) %>%   # NA first, highlighted last
    pull(cell)
  
  # Plot with ordered cells
  p <- DimPlot(
    filtered2,
    reduction = "umap",
    group.by = "highlighted_group",
    cols = highlight_cols,
    na.value = "grey95",
    cells = ordered_cells
  ) +
    ggtitle(paste("UMAP with", group, "Highlighted")) +
    plot_layout(ncol = 1, nrow = 1,
                widths = unit(3, "in"),
                heights = unit(3, "in"))
  
  print(p)
}


```

```{r, dev=c('png', 'pdf'), fig.width=20, fig.height=6}

DimPlot(filtered2, reduction = "umap", 
        split.by = 'group', 
        cols = col_group, group.by = 'group') +
  plot_layout(ncol = 1, nrow = 1,     
                        widths = unit(12, "in"),   
                        heights = unit(3, "in")) 

DimPlot(filtered2, reduction = "umap", 
        split.by = 'group', 
        cols = col_cluster) +
    plot_layout(ncol = 1, nrow = 1,     
                        widths = unit(12, "in"),   
                        heights = unit(3, "in")) 

```

```{r, dev=c('png', 'pdf')}
#| fig-cap: "Bar plot colored by different grouping"
#| layout-ncol: 4

ggplot(filtered2@meta.data, aes(x = seurat_clusters)) +
  geom_bar(aes(fill = Cell_Type), position = "fill") +
  scale_fill_manual(
    values   = col_source,
    name     = "Source"
  ) +
  labs(x = "Cluster", y = "Proportion") +
  force_panelsizes(rows = unit(4, 'inch'), cols = unit(3, 'inch')) 

ggplot(filtered2@meta.data, aes(x = seurat_clusters)) +
  geom_bar(aes(fill = Treatment), position = "fill") +
    scale_fill_manual(
    values   = col_treatment,
    name     = "Treatment"
  ) +
  labs(x = "Cluster", y = "Proportion") +
  force_panelsizes(rows = unit(4, 'inch'), cols = unit(3, 'inch')) 

ggplot(filtered2@meta.data, aes(x = seurat_clusters)) +
  geom_bar(aes(fill = group), position = "fill") +
  scale_fill_manual(
    values   = col_group,
    name     = "Source_Treatment"
  ) +
  labs(x = "Cluster", y = "Proportion") +
  force_panelsizes(rows = unit(4, 'inch'), cols = unit(3, 'inch')) 

ggplot(filtered2@meta.data, aes(x = seurat_clusters)) +
  geom_bar(aes(fill = celltype_merge), position = "fill") +
  scale_fill_manual(
    values   = col_facs,
    na.value = na_grey,
    name     = "FACS Immunophenotype"
  ) +
  labs(x = "Cluster", y = "Proportion") +
  force_panelsizes(rows = unit(4, 'inch'), cols = unit(3, 'inch')) 

# 0: HSC/MHSC
# 7: lymphoid
# 7/8: CB DMSO

```

### AZA-induced Seurat cluster expansion/shrinking
```{r, dev=c('png', 'pdf'), fig.width=8, fig.height=6}

ggplot(filtered2@meta.data, aes(x = group)) +
  geom_bar(aes(fill = seurat_clusters), position = "fill") +
    scale_fill_manual(
    values   = col_cluster,
    na.value = na_grey,
    name     = "Seurat Clusters"
  ) +
  facet_grid(. ~ Cell_Type, scales = 'free_x') +
  labs(x = "Group", y = "Proportion") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  force_panelsizes(rows = unit(4, 'inch'), cols = unit(3, 'inch')) 

props <- filtered2@meta.data %>%
  count(group, seurat_clusters, name = "n", .drop = FALSE) %>%
  group_by(group) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  separate(group, c(NA, 'Cell_Type', 'Treatment'), remove = F)

ggplot(props, aes(x = group, y = prop, group = seurat_clusters, color = seurat_clusters)) +
  geom_line() +
  geom_point(size = 3) +
  geom_text_repel(
    aes(label = scales::percent(prop, accuracy = 0.1)),
    size = 3, show.legend = FALSE
  ) +
  facet_grid(. ~ Cell_Type, scales = 'free_x') +
  labs(x = NULL, y = "Proportion") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  force_panelsizes(rows = unit(4, 'inch'), cols = unit(3, 'inch')) +
  scale_color_manual(
    values   = col_cluster,
    na.value = na_grey,
    name     = "Seurat Clusters"
  ) 

ggplot(props %>%
         mutate(seurat_clusters = factor(seurat_clusters, 
                                         levels = c(0,8,1,3,5,2,6,4,7),
                                         labels = c("HSC/MPP", "cycling/LMPP", "LMPP/Neu1", "Neu1", "Neu1/2", "MEP", "EBM", "EryP", "MLP"))), 
       aes(x = seurat_clusters, y = prop, fill = group)) +
  geom_col(position = position_dodge(width = 0.8, preserve = "single"), 
           width = 0.8) +
  labs(x = "Cell type", y = "Proportion", fill = "Group") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(.~Cell_Type, scales = "free_x", space = "free_x") +
  scale_fill_manual(
    values   = col_group,
    na.value = na_grey,
    name     = "Group"
  )

```


### AZA-induced FACS cluster expansion/shrinking
```{r, dev=c('png', 'pdf'), fig.width=8, fig.height=6}

ggplot(filtered2@meta.data, aes(x = group)) +
  geom_bar(aes(fill = celltype_merge), position = "fill") +
    scale_fill_manual(
    values   = col_facs,
    na.value = na_grey,
    name     = "FACS"
  ) +
  facet_grid(. ~ Cell_Type, scales = 'free_x') +
  labs(x = "Group", y = "Proportion") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  force_panelsizes(rows = unit(4, 'inch'), cols = unit(3, 'inch')) 

props <- filtered2@meta.data %>%
  count(group, celltype_merge, name = "n", .drop = FALSE) %>%
  group_by(group) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  separate(group, c(NA, 'Cell_Type', 'Treatment'), remove = F)

ggplot(props, aes(x = group, y = prop, group = celltype_merge, color = celltype_merge)) +
  geom_line() +
  geom_point(size = 3) +
  geom_text_repel(
    aes(label = scales::percent(prop, accuracy = 0.1)),
    size = 3, show.legend = FALSE
  ) +
  facet_grid(. ~ Cell_Type, scales = 'free_x') +
  labs(x = NULL, y = "Proportion") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  force_panelsizes(rows = unit(4, 'inch'), cols = unit(3, 'inch')) +
  scale_color_manual(
    values   = col_facs,
    na.value = na_grey,
    name     = "FACS"
  ) 

ggplot(props %>%
         filter(!(celltype_merge %in% c('DEAD','other'))), 
       aes(x = celltype_merge, y = prop, fill = group)) +
  geom_col(position = position_dodge(width = 0.8, preserve = "single"), 
           width = 0.8) +
  labs(x = "Cell type", y = "Proportion", fill = "Group") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(.~Cell_Type, scales = "free_x", space = "free_x") +
  scale_fill_manual(
    values   = col_group,
    na.value = na_grey,
    name     = "Group"
  )

```


### Find cluster marker gene
Top 20 markers for all clusters.  
```{r, dev=c('png', 'pdf'), fig.width=15, fig.height=5}

cluster.markers <- FindAllMarkers(filtered2, only.pos = TRUE)

# Create an interactive datatable, show only top 20 (by p value) positive markers in each cluster
datatable(cluster.markers[, c(6,3,4,2,1,5,7)] %>%
            group_by(cluster) %>% 
            arrange(p_val) %>% 
            slice_head(n = 20) %>% 
            ungroup(), 
          filter = "top", caption = 'Top 20 DEG in each cluster',
          options = list(pageLength = 20, autoWidth = TRUE)) %>%
  formatSignif(columns = 2:6, digits = 3)


# selected marker based on pct, logFC and p value
selected_marker <- cluster.markers %>%
  group_by(cluster) %>% 
  arrange(p_val) %>% 
  slice_head(n = 3) %>% 
  ungroup() %>% pull(gene)

DotPlot(filtered2, features = selected_marker %>% unique())  + 
  RotatedAxis() +
  force_panelsizes(rows = unit(3, 'inch'), cols = unit(length(selected_marker)/4, 'inch')) +
  scale_color_viridis(direction = 1) +
  ggtitle('Top 3 markers in each cluster')


selected_marker <- cluster.markers %>%
  group_by(cluster) %>% 
  arrange(p_val) %>% 
  filter(pct.1 > 0.5, pct.2 < 0.2, avg_log2FC > 1) %>%
  slice_head(n = 3) %>% 
  ungroup() %>% pull(gene)

DotPlot(filtered2, features = selected_marker %>% unique())  + 
  RotatedAxis() +
  force_panelsizes(rows = unit(3, 'inch'), cols = unit(length(selected_marker)/4, 'inch')) +
  scale_color_viridis(direction = 1) +
  ggtitle('Top 3 markers in each cluster\n(pct.cluster > 0.5, pct.rest < 0.2, avg_log2FC > 1)')

# show the markers from Dev cell paper
marker <- read.table('data/dev_cell_2022/Cluster_marker.txt', sep = '\t', header = T)

marker <- marker %>% filter(avg_logFC>0 ) %>%
  group_by(cluster) %>%
  arrange(p_val_adj) %>%
  slice_head(n=3) %>% 
  pull(gene) %>%
  unique()

DotPlot(filtered2, features = marker)  + 
  RotatedAxis() +
  force_panelsizes(rows = unit(3, 'inch'), cols = unit(length(marker)/4, 'inch')) +
  scale_color_viridis(direction = 1) +
  ggtitle('Top 3 markers in DevCell subtype from their paper')

# cluster 7: MLP
DotPlot(filtered2, c('CLNK','CCR7','KCNK17')) +
  RotatedAxis() +
  scale_color_viridis(direction = 1) +
  ggtitle('Top 3 markers DevCell MLP')
# cluster 0: HSC/PP
DotPlot(filtered2, c('AVP','CRHBP','DLK1','RPL31','ENO1','RPS3','RPL33','RPS14')) +
  RotatedAxis() +
  scale_color_viridis(direction = 1) +
  ggtitle('Top 3 markers DevCell HSC/MPP1/MPP2')
# cluster 1: LMPP/Neu1
DotPlot(filtered2, c('ENO1','GAPDH','SMIM24','MPO','CDK6','FAM101B')) +
  RotatedAxis() +
  scale_color_viridis(direction = 1) +
  ggtitle('Top 3 markers DevCell LMPP/Neu1')
# cluster 8: cycling LMPP
DotPlot(filtered2, c('ENO1','GAPDH','SMIM24')) +
  RotatedAxis() +
  scale_color_viridis(direction = 1) +
  ggtitle('Top 3 markers DevCell LMPP') # and Histone
# cluster 2: MEP
DotPlot(filtered2, c('CTNNBL1','CNRIP1','SLC40A1')) +
  RotatedAxis() +
  scale_color_viridis(direction = 1) +
  ggtitle('Top 3 markers DevCell MEP')
# cluster 4: EryP
DotPlot(filtered2, c('APOC1','FAM178B','CNRIP1')) +
  RotatedAxis() +
  scale_color_viridis(direction = 1) +
  ggtitle('Top 3 markers DevCell Ery')
# cluster 6: EBM
DotPlot(filtered2, c('IL5RA','PRG2','BACE2')) +
  RotatedAxis() +
  scale_color_viridis(direction = 1) +
  ggtitle('Top 3 markers DevCell EBM')
# cluster 5: Neu1/2
DotPlot(filtered2, c('MPO','CDK6','FAM101B', 'CSTA','SERPINB10','ANKRD18A')) +
  RotatedAxis() +
  scale_color_viridis(direction = 1) +
  ggtitle('Top 3 markers DevCell Neu1/Neu2')
# cluster 3: Neu1
DotPlot(filtered2, c('MPO','CDK6','FAM101B')) +
  RotatedAxis() +
  scale_color_viridis(direction = 1) +
  ggtitle('Top 3 markers DevCell Neu1')

DotPlot(filtered2, features = c('CRHBP','DLK1','RPL31','ENO1','RPS3','RPS14', # HSC/MPP
                                'CTNNBL1','CNRIP1','SLC40A1', # MEP
                                'MPO','CDK6', # Neu1
                                'APOC1','FAM178B',  # Ery
                                'CSTA','SERPINB10','ANKRD18A', # Neu2
                                'IL5RA','PRG2','BACE2', # EBM
                                'CLNK','CCR7','KCNK17', # LMPP/Neu1
                                'GAPDH','SMIM24' # cycling LMPP
                                )) +   
  RotatedAxis() +
  force_panelsizes(rows = unit(3, 'inch'), cols = unit(24/4, 'inch')) +
  scale_color_viridis(direction = 1) +
  ggtitle('Top 3 DevCell markers in each cluster')

```


## Cell type annotation 

### use Dev Cell 2022 data 
```{r, dev=c('png', 'pdf')}
## run azimuth

filtered2 <- RunAzimuth(
  query = filtered2,
  reference = "../PD7151/R/ref_azimuth/" 
)

filtered2$predicted.RNA_clusters <- factor(filtered2$predicted.RNA_clusters, 
                                            levels = names(col_ref))

DimPlot(filtered2, reduction = "umap",  group.by = 'predicted.RNA_clusters',
        cols = col_ref) +
  ggtitle("Annotated to Dev Cell using Azimuth") +
  plot_layout(ncol = 1, nrow = 1,     
              widths = unit(3, "in"),   
              heights = unit(3, "in") )

ggplot(filtered2@meta.data, aes(x = seurat_clusters)) +
  geom_bar(aes(fill = predicted.RNA_clusters), position = "fill") +
  scale_fill_manual(
    values   = col_ref,
    na.value = na_grey,
    name     = "Azimuth Dev Cell"
  ) +
  labs(x = "Cluster", y = "Proportion") +
  force_panelsizes(rows = unit(4, 'inch'), cols = unit(3, 'inch')) 

## compare dev cell vs facs clusters
filtered2@meta.data %>%
  filter(Cell_Type != 'MDSL') %>%
  group_by(Cell_Type, Treatment) %>%
  group_modify(~make_long(.x, celltype_merge, predicted.RNA_clusters)) %>%
  mutate(node = factor(node, 
                       levels = c(
                         rev(names(col_ref)),
                         rev(names(col_facs))) %>% rev() %>% unique() %>% rev() ) 
  ) %>% 
  ggplot(aes(x = x, 
             next_x = next_x, 
             node = node, 
             next_node = next_node,
             fill = node,
             label = node)) +
  geom_sankey(flow.alpha = .6,
              node.color = "gray30") +
  geom_sankey_label(size = 3, color = "white", fill = "gray40") +
  # scale_fill_viridis_d(drop = FALSE) +
  scale_fill_manual(values = c(col_facs, 
                               col_ref
  ))  

saveRDS(filtered2, 'output/filtered2.rds')
filtered2@meta.data %>% write.csv('output/master_filtered.csv')
```


```{r}
sessionInfo()
```
