---
title: "SCIMETAR single cell analysis (Flitered by RPKM) add LINE1, FACS and amplicon data"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
    code-overflow: wrap
    number-sections: true
    number-depth: 3
execute: 
  cache: true
  warning: false
  error: false
---

## Load filtered and annotated seurat object 

```{r, dev=c('png', 'pdf')}
library(matrixStats)
library(ggfortify)
library(Seurat)
library(edgeR)
library(tidyverse)
library(DT)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggpubr)
library(EnhancedVolcano)
library(ggh4x)
library(viridis)
library(patchwork)
library(readxl)
library(RColorBrewer)
library(UpSetR)
library(data.table)
library(ggsankey)
library(ggalluvial)
library(ggrepel)
library(scales)
library(ggsci)
```


```{r}

filtered2 <- readRDS("output/filtered2.rds")

# 2 group genotype
filtered2 <- AddMetaData(filtered2, 
                         metadata = data.frame(genotype_srsf2_2 = str_replace(filtered2$genotype_srsf2, 'mut/mut','mut/wt'),
                                               genotype_tet2a_2 = str_replace(filtered2$genotype_tet2a, 'mut/mut','mut/wt'),
                                               genotype_tet2b_2 = str_replace(filtered2$genotype_tet2b, 'mut/mut','mut/wt')))

# 1615 cells
filtered2_pass <- subset(filtered2, subset = keep_median & dCt > 0)

# harmozine gene name
gencode_gtf <- readRDS('/home/users/allstaff/yan.a/lab_davidson/yan.a/ref/gencode/gencodev44.rds')

genes <- gencode_gtf %>% plyranges::filter(type=='gene') %>%
  plyranges::select(gene_id, gene_name) %>% data.frame()

m <- genes$gene_id %>%
  str_remove("\\..*") %>%
  unique() %>%
  bitr(fromType = "ENSEMBL",
       toType   = c("ENTREZID","SYMBOL"),
       OrgDb    = org.Hs.eg.db)

m1 <- m %>%
  left_join(genes %>% transmute(ENSEMBL = str_remove(gene_id, "\\..*"),
                                gene_name),
            by = "ENSEMBL") %>%
  mutate(match_name = (!is.na(gene_name) & SYMBOL == gene_name)) %>%
  arrange(ENSEMBL, desc(match_name), is.na(ENTREZID)) %>%  # prefer name match, then non-NA
  distinct(ENSEMBL, .keep_all = TRUE) %>%
  dplyr::select(-gene_name, -match_name)

genes <- genes %>% 
  mutate(ENSEMBL = str_remove(gene_id, "\\..*")) %>% 
  left_join(m1, by = 'ENSEMBL')

```

## DEG analysis (separate by source and treatment)

### DEG analysis in each group comparing the top 1/3 vs bottom 1/3 based on LINE1
* Only remove failed RNAseq cells.  
* Cell.1 is top and Cell.2 is bottom.  
* Positive logFC means higher expression in top.
* Use dCt_adj

```{r,fig.width=20, fig.height=15,  dev=c('png', 'pdf')}

deglist2 <- lapply(c("Primary_Patient_DMSO", "Primary_Patient_AZA", 
                            "Primary_CB34_DMSO",   "Primary_CB34_AZA"   ), function(x){
  sub <- subset(filtered2_pass, subset = group == x)
  
  median <- median(sub$dCt_adj)
  tercicle <- quantile(sub$dCt_adj, probs = seq(0, 1, length.out = 4))
  
  sub <- AddMetaData(sub, 
                     metadata = data.frame(Median = ifelse(sub$dCt_adj >= median, 'top','bottom'),
                                           Terciles = ifelse(sub$dCt_adj >= tercicle[3], 'top', 
                                                             ifelse(sub$dCt_adj < tercicle[2], 'bottom', 'other'))))
  deg <- tryCatch(
    {
    FindMarkers(
  object        = sub,
  ident.1       = "top",               # which treatment to test
  ident.2       = "bottom",              # what to compare against, baseline
  group.by      = "Terciles",         # which metadata column defines AZA vs DMSO
  # subset.ident  = "Primary_Patient",   # first subset to only these cells
  only.pos      = FALSE
) %>% rownames_to_column(var = 'gene_name') %>%
  left_join(genes, by = 'gene_name')
  },
  error = function(e) data.frame()
)
})
 
names(deglist2) <- c("Primary_Patient_DMSO", "Primary_Patient_AZA", 
                            "Primary_CB34_DMSO",   "Primary_CB34_AZA"   )

gene.list.grp2 <- lapply(deglist2, function(x) {
                        if (nrow(x) != 0) {
                            x %>% filter(p_val < 0.05) %>% pull(ENTREZID)
                        }
})

fromList(gene.list.grp2) %>%
  UpSetR::upset(nsets = length(gene.list.grp2))

go.grp2 <- gene.list.grp2 %>%
  compareCluster(fun = 'enrichGO', OrgDb = org.Hs.eg.db, ont = 'BP', readable = T) %>%
  clusterProfiler::simplify()

dotplot(go.grp2, label_format = 50) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# cnetplot(go.grp2, foldChange = setNames(deg.pp$avg_log2FC, deg.pp$ENTREZID))

lapply(names(deglist2), function(x) {
  if (nrow(deglist2[[x]]) != 0) {
      deglist2[[x]] %>% mutate(comparison = x) %>%
    filter(p_val < 0.05)
  }
}) %>% rbindlist() %>%
  write.csv(., file = 'output/line1_deg_group2.csv', quote = F, row.names = F)

write.csv(go.grp2@compareClusterResult, 
           file = 'output/line1_deg_group_go2.csv', quote = F, row.names = F)

# split up and down
gene.list.split2 <- c(lapply(deglist2, function(x) {
  x %>% filter(p_val < 0.05, avg_log2FC > 0) %>%
    pull(ENTREZID)
}),  lapply(deglist2, function(x) {
  x %>% filter(p_val < 0.05, avg_log2FC < 0) %>%
    pull(ENTREZID)
}))
names(gene.list.split2) <- paste0(names(gene.list.split2), rep(c('_up', '_down'), each = 4))

fromList(gene.list.split2) %>%
  UpSetR::upset(sets = names(gene.list.split2))

go.split2 <- gene.list.split2 %>%
  compareCluster(fun = 'enrichGO', OrgDb = org.Hs.eg.db, ont = 'BP', readable = T) %>%
  clusterProfiler::simplify()

dotplot(go.split2)
cnetplot(go.split2
         # , foldChange = setNames(deg.pp$avg_log2FC, deg.pp$ENTREZID)
         )

write.csv(go.split2@compareClusterResult, 
           file = 'output/line1_deg_group_go_splitdirection2.csv', quote = F, row.names = F)
```

## DEG analysis (only separate by source)

### DEG analysis in each cluster comparing the top 1/3 vs bottom 1/3 based on LINE1
* Only remove failed RNAseq cells.  
* Cell.1 is top and Cell.2 is bottom.  
* Positive logFC means higher expression in top.
* Use dCt_adj

```{r,fig.width=20, fig.height=15,  dev=c('png', 'pdf')}

deglist.pt2 <- lapply(0:8, function(x){
  sub <- subset(filtered2_pass, subset = seurat_clusters == x & Cell_Type == 'Primary_Patient')
  
  median <- median(sub$dCt_adj)
  tercicle <- quantile(sub$dCt_adj, probs = seq(0, 1, length.out = 4))
  
  sub <- AddMetaData(sub, 
                     metadata = data.frame(Median = ifelse(sub$dCt_adj >= median, 'top','bottom'),
                                           Terciles = ifelse(sub$dCt_adj >= tercicle[3], 'top', 
                                                             ifelse(sub$dCt_adj < tercicle[2], 'bottom', 'other'))))
  deg <- tryCatch(
    {
    FindMarkers(
  object        = sub,
  ident.1       = "top",               # which treatment to test
  ident.2       = "bottom",              # what to compare against, baseline
  group.by      = "Terciles",         # which metadata column defines AZA vs DMSO
  # subset.ident  = "Primary_Patient",   # first subset to only these cells
  only.pos      = FALSE
) %>% rownames_to_column(var = 'gene_name') %>%
  left_join(genes, by = 'gene_name')
  },
  error = function(e) data.frame()
)

})

deglist.cb2 <- lapply(0:8, function(x){
  sub <- subset(filtered2_pass, subset = seurat_clusters == x & Cell_Type == 'Primary_CB34')
  
  median <- median(sub$dCt_adj)
  tercicle <- quantile(sub$dCt_adj, probs = seq(0, 1, length.out = 4))
  
  sub <- AddMetaData(sub, 
                     metadata = data.frame(Median = ifelse(sub$dCt_adj >= median, 'top','bottom'),
                                           Terciles = ifelse(sub$dCt_adj >= tercicle[3], 'top', 
                                                             ifelse(sub$dCt_adj < tercicle[2], 'bottom', 'other'))))
  deg <- tryCatch(
    {
    FindMarkers(
  object        = sub,
  ident.1       = "top",               # which treatment to test
  ident.2       = "bottom",              # what to compare against, baseline
  group.by      = "Terciles",         # which metadata column defines AZA vs DMSO
  # subset.ident  = "Primary_CB34",   # first subset to only these cells
  only.pos      = FALSE
) %>% rownames_to_column(var = 'gene_name') %>%
  left_join(genes, by = 'gene_name')
  },
  error = function(e) data.frame()
)

})
 
names(deglist.pt2) <- paste0('Patient_Cluster', 0:8)
names(deglist.cb2) <- paste0('CB_Cluster', 0:8)

gene.list.src2 <- lapply(c(deglist.pt2, deglist.cb2), function(x) {
  if (nrow(x) != 0) {
      x %>% filter(p_val < 0.05) %>% pull(ENTREZID)
  }
})

fromList(gene.list.src2) %>%
  UpSetR::upset(nsets = 18)

go.src2 <- gene.list.src2 %>%
  compareCluster(fun = 'enrichGO', OrgDb = org.Hs.eg.db, ont = 'BP', readable = T) %>%
  clusterProfiler::simplify()

dotplot(go.src2, label_format = 50) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
# cnetplot(go.src2, foldChange = setNames(deg.pp$avg_log2FC, deg.pp$ENTREZID))


list.all.src2 <- c(deglist.pt2, deglist.cb2) 
lapply(names(list.all.src2), function(x) {
  if (nrow(list.all.src2[[x]]) != 0) {
      list.all.src2[[x]] %>% mutate(comparison = x) %>%
    filter(p_val < 0.05)
  }
}) %>% rbindlist() %>%
  write.csv(., file = 'output/line1_deg_source2.csv', quote = F, row.names = F)

write.csv(go.src2@compareClusterResult, 
           file = 'output/line1_deg_source_go2.csv', quote = F, row.names = F)
```

## DEG analysis (separate by source and treatment)

### DEG analysis in each cluster comparing the top 1/3 vs bottom 1/3 based on LINE1
* Only remove failed RNAseq cells.  
* Cell.1 is top and Cell.2 is bottom.  
* Positive logFC means higher expression in top.
* Use dCt_adj

```{r,fig.width=20, fig.height=15,  dev=c('png', 'pdf')}

Idents(filtered2_pass) <- filtered2_pass$Cell_Type

deglist.pt.dmso2 <- lapply(0:8, function(x){
  sub <- subset(filtered2_pass, subset = seurat_clusters == x & group == 'Primary_Patient_DMSO')
  
  median <- median(sub$dCt_adj)
  tercicle <- quantile(sub$dCt_adj, probs = seq(0, 1, length.out = 4))
  
  sub <- AddMetaData(sub, 
                     metadata = data.frame(Median = ifelse(sub$dCt_adj >= median, 'top','bottom'),
                                           Terciles = ifelse(sub$dCt_adj >= tercicle[3], 'top', 
                                                             ifelse(sub$dCt_adj < tercicle[2], 'bottom', 'other'))))
  deg <- tryCatch(
    {
    FindMarkers(
  object        = sub,
  ident.1       = "top",               # which treatment to test
  ident.2       = "bottom",              # what to compare against, baseline
  group.by      = "Terciles",         # which metadata column defines AZA vs DMSO
  # subset.ident  = "Primary_Patient",   # first subset to only these cells
  only.pos      = FALSE
) %>% rownames_to_column(var = 'gene_name') %>%
  left_join(genes, by = 'gene_name')
  },
  error = function(e) data.frame()
)
})

deglist.pt.aza2 <- lapply(0:8, function(x){
  sub <- subset(filtered2_pass, subset = seurat_clusters == x & group == 'Primary_Patient_AZA')
  
  median <- median(sub$dCt_adj)
  tercicle <- quantile(sub$dCt_adj, probs = seq(0, 1, length.out = 4))
  
  sub <- AddMetaData(sub, 
                     metadata = data.frame(Median = ifelse(sub$dCt_adj >= median, 'top','bottom'),
                                           Terciles = ifelse(sub$dCt_adj >= tercicle[3], 'top', 
                                                             ifelse(sub$dCt_adj < tercicle[2], 'bottom', 'other'))))
  deg <- tryCatch(
    {
    FindMarkers(
  object        = sub,
  ident.1       = "top",               # which treatment to test
  ident.2       = "bottom",              # what to compare against, baseline
  group.by      = "Terciles",         # which metadata column defines AZA vs DMSO
  # subset.ident  = "Primary_Patient",   # first subset to only these cells
  only.pos      = FALSE
) %>% rownames_to_column(var = 'gene_name') %>%
  left_join(genes, by = 'gene_name')
  },
  error = function(e) data.frame()
)
})

deglist.cb.dmso2 <- lapply(0:8, function(x){
  sub <- subset(filtered2_pass, subset = seurat_clusters == x & group == 'Primary_CB34_DMSO')
  
  median <- median(sub$dCt_adj)
  tercicle <- quantile(sub$dCt_adj, probs = seq(0, 1, length.out = 4))
  
  sub <- AddMetaData(sub, 
                     metadata = data.frame(Median = ifelse(sub$dCt_adj >= median, 'top','bottom'),
                                           Terciles = ifelse(sub$dCt_adj >= tercicle[3], 'top', 
                                                             ifelse(sub$dCt_adj < tercicle[2], 'bottom', 'other'))))
  deg <- tryCatch(
    {
    FindMarkers(
  object        = sub,
  ident.1       = "top",               # which treatment to test
  ident.2       = "bottom",              # what to compare against, baseline
  group.by      = "Terciles",         # which metadata column defines AZA vs DMSO
  # subset.ident  = "Primary_Patient",   # first subset to only these cells
  only.pos      = FALSE
) %>% rownames_to_column(var = 'gene_name') %>%
  left_join(genes, by = 'gene_name')
  },
  error = function(e) data.frame()
)
})

deglist.cb.aza2 <- lapply(0:8, function(x){
  sub <- subset(filtered2_pass, subset = seurat_clusters == x & group == 'Primary_CB34_AZA')
  
  median <- median(sub$dCt_adj)
  tercicle <- quantile(sub$dCt_adj, probs = seq(0, 1, length.out = 4))
  
  sub <- AddMetaData(sub, 
                     metadata = data.frame(Median = ifelse(sub$dCt_adj >= median, 'top','bottom'),
                                           Terciles = ifelse(sub$dCt_adj >= tercicle[3], 'top', 
                                                             ifelse(sub$dCt_adj < tercicle[2], 'bottom', 'other'))))
  deg <- tryCatch(
    {
    FindMarkers(
  object        = sub,
  ident.1       = "top",               # which treatment to test
  ident.2       = "bottom",              # what to compare against, baseline
  group.by      = "Terciles",         # which metadata column defines AZA vs DMSO
  # subset.ident  = "Primary_Patient",   # first subset to only these cells
  only.pos      = FALSE
) %>% rownames_to_column(var = 'gene_name') %>%
  left_join(genes, by = 'gene_name')
  },
  error = function(e) data.frame()
)
})
 
names(deglist.pt.aza2) <- paste0('Patient_AZA_Cluster', 0:8)
names(deglist.pt.dmso2) <- paste0('Patient_DMSO_Cluster', 0:8)

names(deglist.cb.aza2) <- paste0('CB_AZA_Cluster', 0:8)
names(deglist.cb.dmso2) <- paste0('CB_DMSO_Cluster', 0:8)

gene.list.st2 <- lapply(c(deglist.pt.aza2, deglist.pt.dmso2, 
                      deglist.cb.aza2, deglist.cb.dmso2), function(x) {
                        if (nrow(x) != 0) {
                            x %>% filter(p_val < 0.05) %>% pull(ENTREZID)
                        }
})

fromList(gene.list.st2) %>%
  UpSetR::upset(nsets = length(gene.list.st2))

go.st2 <- gene.list.st2 %>%
  compareCluster(fun = 'enrichGO', OrgDb = org.Hs.eg.db, ont = 'BP', readable = T) %>%
  clusterProfiler::simplify()

dotplot(go.st2, label_format = 50) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# cnetplot(go.st2, foldChange = setNames(deg.pp$avg_log2FC, deg.pp$ENTREZID))

list.all.st2 <- c(deglist.pt.aza2, deglist.pt.dmso2, deglist.cb.aza2, deglist.cb.dmso2) 
lapply(names(list.all.st2), function(x) {
  if (nrow(list.all.st2[[x]]) != 0) {
      list.all.st2[[x]] %>% mutate(comparison = x) %>%
    filter(p_val < 0.05)
  }
}) %>% rbindlist() %>%
  write.csv(., file = 'output/line1_deg_source_treatment2.csv', quote = F, row.names = F)

write.csv(go.st2@compareClusterResult, 
           file = 'output/line1_deg_source_treatment_go2.csv', quote = F, row.names = F)
```
