---
title: "QC"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
    code-overflow: wrap
execute: 
  cache: true
  warning: false
  error: false
---


## results from zUMI parameter

Use zUMI default.
```
if [ ! -f $out_dir/*_Aligned.sortedByCoord.out.bam ]; then
    # use new 2pass mode of STAR
    STAR --twopassMode Basic \
        --genomeDir $GENOME_DIR \
        --readFilesIn ${R1} ${R2} \
        --runThreadN ${ncpu} \
        --sjdbGTFfile ${GTF} \
        --sjdbOverhang ${sjdbOverhang} \
        --readFilesCommand zcat \
        --outSAMmultNmax 1 \
        --outFilterMultimapNmax 50 \
        --outSAMtype BAM SortedByCoordinate \
        --outSAMattrRGline ID:${sample_ID} LB:library PL:illumina PU:machine SM:${sample_ID} \
        --outFileNamePrefix $out_dir/${sample_ID}_ 

    rm -rf $out_dir/*__STAR*    
	samtools index $out_dir/${sample_ID}_Aligned.sortedByCoord.out.bam
fi
```
```{r, fig.width=10, dev=c('png', 'pdf')}
library(data.table)
library(tidyverse)

## star summary

star2 <- fread('../align_zumi/multiqc_data/star_alignment_plot.txt')

colnames(star2) <- c('Sample',
  "Uniquely_mapped",
  "Mapped_to_multiple_loci",
  "Mapped_to_too_many_loci",
  #"Unmapped_too_many_mismatches",
  "Unmapped_too_short",
  "Unmapped_other"
)

align_col <- pal_npg(palette = c("nrc"), alpha = 1)(6)
names(align_col) <- colnames(star)[-1]

star2 <- star2 %>%
  # 1) Create a Total_Count column
  mutate(Total_Count = rowSums(across(
    c(Uniquely_mapped,
      Mapped_to_multiple_loci,
      Mapped_to_too_many_loci,
      #Unmapped_too_many_mismatches,
      Unmapped_too_short,
      Unmapped_other))),
  Sample = sub("^((?:[^_]+_){3}[^_]+).*", "\\1", Sample)) %>%
  # 2) Create percentage columns for each category
  mutate(across(
    c(Uniquely_mapped,
      Mapped_to_multiple_loci,
      Mapped_to_too_many_loci,
      #Unmapped_too_many_mismatches,
      Unmapped_too_short,
      Unmapped_other),
    ~ (. / Total_Count) * 100,
    .names = "{.col}_pct"
  )) %>%
  separate(Sample, c('Plate', 'Source', 'Treatment', 'Well'), remove = F) %>%
  mutate(is.empty = str_detect(Well, '87|91|94'))
```

### Summary of STAR aligner
Number of aligned reads in each category
```{r, fig.width=10, dev=c('png', 'pdf')}
star2[, c(1:11, 17)] %>%
  pivot_longer(6:11) %>%
  mutate(name = factor(name, levels = c('Total_Count', "Uniquely_mapped",
                                        "Mapped_to_multiple_loci", "Mapped_to_too_many_loci",
                                        "Unmapped_too_short", # "Unmapped_too_many_mismatches",
                                        "Unmapped_other"))) %>%
  ggplot(aes(x = Plate, y = value, fill = is.empty, color = is.empty)) +
  geom_violin(width = 1.4, alpha = 0.5, position = position_dodge(width = 1)) +
  geom_boxplot(width = 0.1, alpha = 0.5, position = position_dodge(width = 1)) +
  facet_wrap(.~name, scales = 'free_y', ncol = 1) +
  ylab('Number of reads') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

star2[, c(1:10, 17)] %>%
  pivot_longer(6:10) %>%
  mutate(name = factor(name, levels = rev(c("Uniquely_mapped",
                                        "Mapped_to_multiple_loci", "Mapped_to_too_many_loci",
                                        "Unmapped_too_short", # "Unmapped_too_many_mismatches",
                                        "Unmapped_other")))) %>%
  ggplot(aes(x = Plate, y = value, fill = name
  )) +
  geom_col(position = 'stack') +
  ylab('Number of reads') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_manual(values = align_col)


star2[, c(1:10, 17)] %>%
  pivot_longer(6:10) %>%
  mutate(name = factor(name, levels = rev(c("Uniquely_mapped",
                                        "Mapped_to_multiple_loci", "Mapped_to_too_many_loci",
                                        "Unmapped_too_short", # "Unmapped_too_many_mismatches",
                                        "Unmapped_other")))) %>%
  ggplot(aes(x = Plate, y = value, fill = name
  )) +
  geom_col(position = 'fill') +
  ylab('Proportion of reads') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_manual(values = align_col)
```


```
GTF="/vast/projects/lab_davidson/yan.a/ref/gencode/gencode.v44.annotation.gtf"

ncpu=48

# 5. featurecount with deduplicated bam

featureCounts -a $GTF -o unstranded_counts.txt -p -T $ncpu  ../align_zumi/*/*_Aligned.sortedByCoord.out.bam
```
This default setting means
```
||             Output file : unstranded_counts.txt                            ||
||                 Summary : unstranded_counts.txt.summary                    ||
||              Paired-end : yes                                              ||
||        Count read pairs : no                                               ||
||              Annotation : gencode.v44.annotation.gtf (GTF)                 ||
||      Dir for temp files : ./                                               ||
||                                                                            ||
||                 Threads : 48                                               ||
||                   Level : meta-feature level                               ||
||      Multimapping reads : not counted                                      ||
|| Multi-overlapping reads : not counted                                      ||
||   Min overlapping bases : 1                                                ||
```
Because the BAM files was not mark duplicates, so cannot see duplicate reads. I think the duplicate removal will be different from bulk RNAseq which simply look at read mapping coordinates.  
```{r, fig.width=10, dev=c('png', 'pdf')}
## featurecount

fc2 <- fread('../count_zumi/multiqc_data/featureCounts_assignment_plot.txt')

colnames(fc2) <- c('Sample',
                    "Assigned",
                    "Unassigned_Multi_Mapping",
                    "Unassigned_No_Features",
                    "Unassigned_Ambiguity")

fc_col <- pal_npg(palette = c("nrc"), alpha = 1)(4)
names(fc_col) <- colnames(fc)[-1]

fc2 <- fc2 %>%
  # 1) Create a Total_Count column
  mutate(Total_Count = rowSums(across(
    c(Assigned,
      Unassigned_Multi_Mapping,
      Unassigned_No_Features,
      Unassigned_Ambiguity))),
    Sample = sub("^((?:[^_]+_){3}[^_]+).*", "\\1", Sample)) %>%
  # 2) Create percentage columns for each category
  mutate(across(
    c(Assigned,
      Unassigned_Multi_Mapping,
      Unassigned_No_Features,
      Unassigned_Ambiguity),
    ~ (. / Total_Count) * 100,
    .names = "{.col}_pct"
  )) %>%
  separate(Sample, c('Plate', 'Source', 'Treatment', 'Well'), remove = F) %>%
  mutate(is.empty = str_detect(Well, '87|91|94'))

```


### Summary of featurecount
Number of counted reads in each category
```{r, fig.width=10, dev=c('png', 'pdf')}
fc2[, c(1:10, 15)] %>%
  pivot_longer(6:10) %>%
  mutate(name = factor(name, levels = c('Total_Count', "Assigned",
                                        "Unassigned_Multi_Mapping",
                                        "Unassigned_No_Features",
                                        "Unassigned_Ambiguity"))) %>%
  ggplot(aes(x = Plate, y = value, fill = is.empty, color = is.empty)) +
  geom_violin(width = 1.4, alpha = 0.5, position = position_dodge(width = 1)) +
  geom_boxplot(width = 0.1, alpha = 0.5, position = position_dodge(width = 1)) +
  facet_wrap(.~name, scales = 'free_y', ncol = 1) +
  ylab('Number of reads') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

fc2[, c(1:10, 15)] %>%
  pivot_longer(6:9) %>%
  mutate(name = factor(name, levels = rev(c("Assigned",
                                        "Unassigned_Multi_Mapping",
                                        "Unassigned_No_Features",
                                        "Unassigned_Ambiguity")))) %>%
  ggplot(aes(x = Plate, y = value, fill = name
  )) +
  geom_col(position = 'stack') +
  ylab('Number of reads') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_manual(values = fc_col)

fc2[, c(1:10, 15)] %>%
  pivot_longer(6:9) %>%
  mutate(name = factor(name, levels = rev(c("Assigned",
                                        "Unassigned_Multi_Mapping",
                                        "Unassigned_No_Features",
                                        "Unassigned_Ambiguity")))) %>%
  ggplot(aes(x = Plate, y = value, fill = name
  )) +
  geom_col(position = 'fill') +
  ylab('Proportion of reads') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_manual(values = fc_col)

```
